---
title: "Getting connected and finding the tables"
subtitle: "Stage one: from an empty R session to a table you can query"
execute:
  freeze: true
---

Two connections, not one, and the reason is a bug rather than a preference. This page opens both, finds the tables, and runs the one check that has to happen before anything else: that geometry survives the trip into your session.

```{r}
#| label: mask
#| include: false
# This site is published, so no printed output may carry the catalog this
# example runs against. Every query below uses the real one; only what is
# shown is masked. assets/redact.lua does the same at pandoc time, but that
# runs after the freeze cache is written and _freeze/ is tracked, so the
# masking has to happen here too.
source(here::here("R/setup.R"))
wq_mask_output()
```

```{r}
#| label: setup
#| message: false
library(DBI)
library(dplyr)
library(dbplyr)
library(tibble)
library(purrr)
library(glue)
library(cli)
library(sf)
```

Where the tables live comes from the environment rather than from anything typed here, so pointing this code at your own tables means changing two variables and nothing else.

```{r}
#| label: location
catalog <- Sys.getenv("DATABRICKS_CATALOG")
schema  <- Sys.getenv("DATABRICKS_SCHEMA")
```

## Opening the connection

For ordinary queries, `odbc::databricks()` against a SQL warehouse is the default and is fine. It is what the next two pages use.

```{r}
#| label: connect-odbc
con <- dbConnect(
  odbc::databricks(),
  httpPath = Sys.getenv("DATABRICKS_PATH"),
  # Load-bearing, not defensive. Left at its default,
  # an encoded geometry is silently clipped at 1,023
  # characters. 65535 is deliberately not round: a
  # round value renders as 1e+05 and is dropped.
  DefaultStringColumnLength = 65535
)
```

Geometry is involved later in this example, and that forces a second connection. `odbc` does not return a `BINARY` column intact, and it does not tell you: a polygon comes back as a fraction of itself, with no error and no warning. Geometry is stored as `BINARY`, so the ODBC path cannot carry it, and [Connect your R session to the data](../howdoi/connect.qmd) explains why this is a constraint rather than a choice.

```{r}
#| label: connect-brickster
acon <- dbConnect(
  brickster::DatabricksSQL(),
  warehouse_id = basename(Sys.getenv("DATABRICKS_PATH"))
)
```

Both point at the same warehouse. The difference is entirely in what survives the wire.

## Finding the tables

Names in Unity Catalog have three parts: catalog, schema, table. The three-part name is not bureaucracy, it is how permission is granted, which is why you can sometimes see a catalog and not its contents.

Rather than jumping to a table name you already know, the honest starting condition is not knowing where anything lives. Browsing costs nothing: these calls ask for metadata and move no data at all.

Name the catalog and schema every time. `dbListTables(con)` without them is not an error and does not warn: it lists whatever the connection happens to default to, which on a shared warehouse is somebody else's schema. A list of unfamiliar table names is easy to read as "my grant has not come through yet" when it means "you are looking in the wrong place".

```{r}
#| label: browse
dbListTables(con, catalog_name = catalog, schema_name = schema)
```

A three-part name is what every query below needs, so it is worth building once rather than pasting together each time.

```{r}
#| label: naming
# in_catalog() is dbplyr's own way to build the three-part
# name. It quotes each part separately, so a catalog or
# table whose name needs escaping still works.
readings_name <- in_catalog(catalog, schema, "hydrology_readings")

readings_name
```

## What the tables look like

Here is the first place this stops resembling a laptop workflow. `tbl()` gives you something you can inspect, count and query, and none of that loads the table.

```{r}
#| label: lazy-table
readings <- tbl(con, readings_name)

readings |>
  glimpse()
```

That listed the columns and their types, and the data is still on Databricks. `glimpse()` needs no `collect()`, because a column listing is metadata. The object is a query, not a table: it describes what to fetch rather than holding anything.

`Rows: ??` is the honest answer rather than a gap. `dbplyr` has not counted them, because counting is a query nobody asked for.

Counting proves the point, because the count runs on the server and only the answer comes back.

```{r}
#| label: row-count
readings |>
  summarise(
    rows = n(),
    first_day = min(date),
    last_day = max(date)
  ) |>
  collect()
```

Thirty-two million rows, and what arrived in your session is three numbers. This is the whole idea the next page is built on.

The other tables are small by comparison, and their sizes are worth knowing now because they decide which ones can simply be pulled across whole.

```{r}
#| label: table-sizes
# as.numeric() is not decoration: the driver returns
# integer64, and map_dbl() would otherwise reinterpret
# its bits rather than convert them.
count_rows <- function(table) {
  tbl(con, in_catalog(catalog, schema, table)) |>
    count() |>
    pull(n) |>
    as.numeric()
}

tibble(
  table = c(
    "hydrology_stations",
    "hydrology_measures",
    "wfd_catchments",
    "storm_overflows",
    "bathing_water_sites"
  )
) |>
  mutate(rows = map_dbl(table, count_rows))
```

## Checking the geometry column

Do this now, not after a spatial join returns nothing. It costs one query, and it is the cheapest possible place to catch a failure that otherwise surfaces much later as a wrong answer.

The check compares two numbers: the length the server says a value has, and the length that actually arrives in R. If they differ, the value was truncated in transit.

```{r}
#| label: truncation-check
longest_five <- tbl(con, in_catalog(catalog, schema, "wfd_catchments")) |>
  filter(!is.na(geom_wkb)) |>
  # base64() and length() both run on the server, so
  # `declared` is the length before the trip.
  transmute(
    g = base64(geom_wkb),
    declared = length(base64(geom_wkb))
  ) |>
  arrange(desc(declared)) |>
  head(5)

longest_five |>
  collect() |>
  transmute(
    declared,
    arrived = nchar(g),
    intact = declared == arrived
  )
```

These arrive intact because the connection above sets `DefaultStringColumnLength` beyond the longest encoded value. Without that setting the same query silently returns 1,023 characters, and the decode that follows fails with an error that reads like corrupt input rather than truncation.

That covers text. Binary is the harder half, and it is not fixable by any driver setting, which is why the second connection exists.

```{r}
#| label: binary-check
#| message: false
bytes_back <- function(connection, n) {
  glue("SELECT cast(repeat('x', {n}) as binary) AS b") |>
    dbGetQuery(conn = connection) |>
    pull(b) |>
    pluck(1) |>
    length()
}

tibble(asked = c(512, 1024, 1025, 2000, 5000)) |>
  mutate(
    odbc = map_dbl(asked, \(n) bytes_back(con, n)),
    brickster = map_dbl(asked, \(n) bytes_back(acon, n))
  )
```

Read the middle column. A 1,025-byte value comes back as a single byte, and nothing about that is an error: the query succeeded and returned a value of the wrong length. The `brickster` column is what correct looks like.

## Proving the geometry

Arriving at full length is necessary but not sufficient. A geometry that parses can still be the wrong geometry, so the useful check is against something recorded independently: the catchment table carries its own area column, computed before any of this transport happened.

```{r}
#| label: area-check
catchments <- tbl(con, in_catalog(catalog, schema, "wfd_catchments")) |>
  filter(!is.na(geom_wkb), !is.na(water_body_area_m2)) |>
  transmute(
    g = base64(geom_wkb),
    declared = water_body_area_m2
  ) |>
  head(200) |>
  collect()

# Two steps, because geometry is encoded server-side to
# survive the trip: it arrives as text, and has to be
# decoded before sf will look at it. The CRS must be
# supplied, because WKB carries no SRID of its own.
geom <- catchments$g |>
  map(jsonlite::base64_dec) |>
  structure(class = "WKB") |>
  st_as_sfc(EWKB = FALSE, crs = 27700)

declared <- sum(catchments$declared)
measured <- sum(as.numeric(st_area(geom)))
err <- abs(measured - declared) / declared

cli_alert_info(
  "EPSG:{st_crs(geom)$epsg}, {length(geom)} polygons, \\
   area differs from declared by {round(100 * err, 4)}%"
)
```

A difference of thousandths of a percent means the geometry is intact and correctly projected. This is the check worth repeating whenever geometry crosses a boundary, and [When the simulation carries the geometry](bootstrap.qmd) uses it again for exactly that reason.

## Configuring more than one compute target

This example uses a warehouse for queries and, later, a cluster for distributed work. Naming both invites a specific silent failure: `config::get()` treats a profile name it does not recognise as absent and falls back to the default, so a typo runs your job on the wrong compute and returns results that look entirely ordinary.

The fix is to validate the name against the file's own keys rather than trusting the lookup.

```{r}
#| label: profile-guard
#| error: true
cluster_id <- function(profile = "default",
                       file = "config.yml") {
  known <- names(yaml::yaml.load_file(file))

  if (!profile %in% known) {
    cli_abort(c(
      "Unknown profile {.val {profile}}.",
      i = "{.file {file}} defines: {.val {known}}.",
      x = "config::get() would have fallen back to
           {.val default} and given you the wrong compute."
    ))
  }

  config::get("cluster_id", config = profile, file = file)
}

# A real profile resolves. The id itself is not printed:
# it names one cluster in one workspace, and this page
# is published.
nzchar(cluster_id("multinode"))

# A typo must not silently succeed.
cluster_id("multinod")
```

Storing the warehouse as a bare identifier rather than as a connection path is the related habit. Two clients want two different forms of the same value, and keeping the shorter one means each composes what it needs instead of taking the longer one apart.

```{r}
#| label: cleanup
dbDisconnect(con)
dbDisconnect(acon)
```

## Next

[Exploring and reducing the big table](reducing.qmd).
