---
title: "Exploring and reducing the big table"
subtitle: "Stage two: getting from tens of millions of rows to something you can hold"
execute:
  freeze: true
---

::: {.callout-note appearance="simple"}

**Draft.** The code runs and the argument holds; the prose has not been copy-edited.

:::

Thirty-two million rows go in and about a thousand come back. The whole of this page is about where that reduction happens, because it decides whether the analysis is possible at all.

```{r}
#| label: setup
#| message: false
library(DBI)
library(dplyr)
library(dbplyr)
library(tibble)
library(glue)
library(cli)

catalog <- Sys.getenv("DATABRICKS_CATALOG")
schema <- Sys.getenv("DATABRICKS_SCHEMA")

con <- dbConnect(
  odbc::databricks(),
  httpPath = Sys.getenv("DATABRICKS_PATH"),
  DefaultStringColumnLength = 65535
)

tbl_name <- function(table) {
  glue("{catalog}.{schema}.{table}")
}

wq_tbl <- function(table) {
  tbl(con, I(tbl_name(table)))
}
```

## Looking before reducing

Everything in this section is ordinary `dplyr`. None of it loads the table, and you can tell which by where `collect()` is: there isn't one yet.

```{r}
#| label: lazy
readings <- wq_tbl("hydrology_readings")
measures <- wq_tbl("hydrology_measures")
```

Start with the shape of the thing. This runs on the warehouse and returns three numbers.

```{r}
#| label: extent
readings |>
  summarise(
    rows = n(),
    first_day = min(date),
    last_day = max(date)
  ) |>
  collect()
```

Ten years of daily data. Before trusting any of it, look at what the quality flag says, because a mean over unchecked values is not the mean you want.

```{r}
#| label: quality
readings |>
  count(quality, sort = TRUE) |>
  collect()
```

About four fifths is `Good`. The rest is not junk, but it is not comparable either, and filtering to `Good` is a decision this analysis makes explicitly rather than by accident.

The readings table does not say what it is measuring. That lives on the measures table, which is small.

```{r}
#| label: parameters
measures |>
  count(parameter_name, period_name, sort = TRUE) |>
  head(8) |>
  collect()
```

## A step that has to move

The analysis wants daily mean flow per station. The obvious way to write that is to filter the readings first and join afterwards, which is the wrong order:

```r
readings |>
  filter(quality == "Good") |>
  inner_join(measures, by = join_by(measure == notation)) |>
  filter(parameter == "flow", period == 86400L)
```

That is not wrong in its answer, it is wrong in where the work lands. The join runs against every `Good` row in the table before anything narrows it to flow. Selecting the measures first means the join has roughly a thousand rows on one side instead of twenty-six million.

The rule generalises: **filter the small side before the join, not the large side after it.**

## Reaching a function `dbplyr` will not translate

Picking daily mean flow needs a match on the end of `value_statistic`, and the natural R way to write that does not survive the trip:

```r
filter(measures, grepl("mean", value_statistic))
```

`dbplyr` translates a great deal, and `grepl()` is not part of it. The query reaches the server with `grepl` still in it and fails there, not here, with `UNRESOLVED_ROUTINE`. That is the signature of a translation gap rather than a mistake in your logic.

The escape hatch is `dplyr::sql()`, which passes a fragment through untouched. You are writing Spark SQL at that point, so the dialect is theirs and not R's.

```{r}
#| label: flow-measures
flow_measures <- measures |>
  filter(
    parameter == "flow",
    period == 86400L,
    sql("value_statistic LIKE '%/mean'")
  )

flow_measures |>
  summarise(measures = n()) |>
  collect()
```

## Where the line falls in this pipeline

Here is the whole reduction, with the one line that matters marked. Everything above `collect()` runs on Databricks; everything below runs in your session.

```{r}
#| label: reduce
per_station <- readings |>
  inner_join(
    flow_measures,
    by = join_by(measure == notation)
  ) |>
  filter(
    quality == "Good",
    !is.na(value),
    value >= 0
  ) |>
  summarise(
    .by = station_id,
    n_days = n(),
    mean_flow = mean(value, na.rm = TRUE),
    sd_flow = sd(value, na.rm = TRUE)
  ) |>
  filter(n_days >= 1000) |>
  # <== everything above this line ran on Databricks
  collect() |>
  mutate(cv = sd_flow / mean_flow)

per_station
```

The `mutate()` sits below the line deliberately. It could run either side, and putting it in the session keeps the SQL to the part that has to be SQL.

`n_days >= 1000` drops stations with less than about three years of record. That is a judgement about what a variability estimate is worth, not a technical constraint.

## What came back

```{r}
#| label: reduction
cli_alert_info(
  "{nrow(per_station)} stations, \\
   {round(as.numeric(object.size(per_station)) / 1024)} KB"
)
```

Thirty-two and a half million rows became about a thousand, and a few hundred kilobytes. Everything after this page is ordinary R on an ordinary data frame.

The coefficient of variation is the number the rest of the example uses. A station with a CV near zero has a steady flow; a high CV means a river that is mostly low and occasionally in flood.

```{r}
#| label: cv
per_station |>
  summarise(
    median_cv = median(cv, na.rm = TRUE),
    min_cv = min(cv, na.rm = TRUE),
    max_cv = max(cv, na.rm = TRUE)
  )
```

## Checking the reduction against the server

The count came back as a single number, so it is worth confirming that number was computed over what you think it was.

```{r}
#| label: crosscheck
server_side <- readings |>
  inner_join(
    flow_measures,
    by = join_by(measure == notation)
  ) |>
  filter(quality == "Good", !is.na(value), value >= 0) |>
  summarise(
    rows = n(),
    stations = n_distinct(station_id)
  ) |>
  collect()

server_side
```

The station count here is higher than the thousand-odd that came back, and that difference is the `n_days >= 1000` filter rather than anything going wrong.

```{r}
#| label: save
saveRDS(per_station, "per_station.rds")
```

```{r}
#| label: cleanup
dbDisconnect(con)
```

## Next

[The spatial join](spatial.qmd).

---

This page rests on: the readings table holds about 32.5 million rows spanning 2015 to 2024; roughly 80% of readings carry quality `Good`; `grepl()` has no Spark SQL translation and fails server-side with `UNRESOLVED_ROUTINE`, so `dplyr::sql()` is needed to match on `value_statistic`; filtering the measures table before the join rather than the readings table after it is what keeps the join small; the reduction yields about a thousand stations and a few hundred kilobytes.
