---
title: "Work with a table that will not fit in your session"
subtitle: "Where each step of your pipeline runs, and how much data comes back"
---

## One line decides where the work happens

`collect()` is the line. Everything you write above it is translated into SQL and runs on Databricks; everything below it runs in your R session, on whatever `collect()` returned. Nothing moves until you call it.

That single sentence is the whole page. Everything else is a consequence of it.

```{r}
#| label: mask
#| include: false
# This site is public, so no printed output may carry the catalog this page
# happens to run against. Only what is *shown* is masked; every query below
# runs against the real tables. See R/setup.R.
source(here::here("R/setup.R"))
wq_mask_output()
```

```{r}
#| label: setup
#| message: false
library(DBI)
library(dplyr)
library(dbplyr)
library(tibble)
library(glue)
library(cli)

catalog <- Sys.getenv("DATABRICKS_CATALOG")
schema <- Sys.getenv("DATABRICKS_SCHEMA")

con <- dbConnect(
  odbc::databricks(),
  httpPath = Sys.getenv("DATABRICKS_PATH"),
  DefaultStringColumnLength = 65535
)

# in_catalog() builds the three-part name and quotes each part, so a
# name needing escaping still works. It is dbplyr's own idiom for this
# and beats pasting the parts together with glue().
readings <- tbl(con, in_catalog(catalog, schema, "hydrology_readings"))
```

`readings` is a query, not a table. `glimpse()` is the quickest way to see its shape, and it needs no `collect()`: the columns and their types come back from the server, and a few values with them.

```{r}
#| label: glimpse
readings |>
  glimpse()
```

Read `Rows: ??` as the honest answer rather than a gap. `dbplyr` does not know how many rows there are, because counting them is a query it has not been asked to run. Everything else on that listing is metadata, which is why it costs nothing.

Printing the object has the same shape of trap, and is worth seeing once:

```{r}
#| label: preview
readings |>
  select(date, value, quality) |>
  head(5)
```

`?? x 3` is `dbplyr` telling you it does not know the row count and has not asked. Ask, and the counting happens on the server:

```{r}
#| label: rowcount
readings |>
  summarise(rows = n()) |>
  collect()
```

### Seeing the line in your own pipeline

Do not take the site's word for where the boundary falls in your code. `show_query()` prints the SQL `dbplyr` will send, so you can read exactly what the server is being asked to do:

```{r}
#| label: showquery
readings |>
  filter(quality == "Good", value >= 0) |>
  summarise(
    .by = measure,
    n_days = n(),
    mean_value = mean(value, na.rm = TRUE)
  ) |>
  show_query()
```

The filter, the grouping and both aggregates are all in the SQL, so all of them run on Databricks. Anything you write after `collect()` will not appear here, and that absence is how you tell the two halves apart. Run `show_query()` on your own pipeline before you run the pipeline, especially when you are about to leave it going while you make coffee.

## What "too big" means here

The readings table is 32,540,721 rows, and collecting the whole thing would put a couple of gigabytes in your R session. On the 32 GB session this guide was written against, that fits. So this table is not, strictly, too big.

It is used here anyway, because the technique is visible on it and because the next table will not fit. The reduction below takes 32.5 million rows down to about a thousand, and that ratio does not change when the input is a hundred times larger. Learn the shape on something you could have collected, and it keeps working on something you could not.

Worth saying plainly: if your data does fit in your laptop's memory and nobody is dictating where it lives, working locally is faster to set up and better for your science. This page is for when it does not, or when someone else decided.

## Three things can happen at that line

Every step in your pipeline is one of three cases, and which one it is depends on two questions. Does Databricks have the function? And does the step produce one row out per row in, or fewer?

### Shrink first

Only R has the function you need, but the steps before it collapse the data. This is the ordinary case and the one to reach for first: do the reduction server-side, then cross with something small.

The analysis wants per-station flow variability. Choosing the right measures narrows one side of the join to about a thousand rows before it touches the readings at all:

```{r}
#| label: measures
measures <- tbl(con, in_catalog(catalog, schema, "hydrology_measures"))

flow_measures <- measures |>
  filter(
    parameter == "flow",
    period == 86400L,
    sql("value_statistic LIKE '%/mean'")
  )

flow_measures |>
  summarise(measures = n()) |>
  collect()
```

Then the reduction itself, with the boundary marked:

```{r}
#| label: reduction
per_station <- readings |>
  inner_join(flow_measures, by = join_by(measure == notation)) |>
  filter(quality == "Good", !is.na(value), value >= 0) |>
  summarise(
    .by = station_id,
    n_days = n(),
    mean_flow = mean(value, na.rm = TRUE),
    sd_flow = sd(value, na.rm = TRUE)
  ) |>
  filter(n_days >= 1000) |>
  # <== everything above this line ran on Databricks
  collect() |>
  mutate(cv = sd_flow / mean_flow)

cli_alert_info(
  "{nrow(per_station)} rows, \\
   {round(as.numeric(object.size(per_station)) / 1024)} KB"
)
```

The server read 3,239,134 rows to produce those 1,029, and none of the three million crossed:

```{r}
#| label: scanned
readings |>
  inner_join(flow_measures, by = join_by(measure == notation)) |>
  filter(quality == "Good", !is.na(value), value >= 0) |>
  summarise(rows = n()) |>
  collect()
```

Three orders of magnitude, from millions to a thousand, and 138 KB in your session. That is the ordinary outcome. Note where `mutate(cv = ...)` sits: it could have run on either side, and putting it below the line keeps the SQL to the part that has to be SQL.

The other thing to notice is the join order. Filtering `measures` down to 1,103 rows *before* the join is what keeps the join small. Filtering the readings first and joining afterwards gives the same answer and makes the server join against every `Good` row in the table. Filter the small side before the join, not the large side after it.

### Stay server-side

You need one row out per row in, and Databricks has the function. Then nothing has to cross at all until the very end, and the intermediate table never exists in your session.

Flagging outliers against each measure's own distribution is one row out per row in, and every piece of it translates:

```{r}
#| label: zscore
readings |>
  filter(quality == "Good", value > 0) |>
  mutate(
    .by = measure,
    z = (value - mean(value, na.rm = TRUE)) / sd(value, na.rm = TRUE)
  ) |>
  filter(z > 5) |>
  summarise(flagged = n()) |>
  collect()
```

The windowed mean and standard deviation become a Spark window function. The 24.7 million rows the `z` column is computed over never leave the warehouse, and one number comes back. If you wanted the flagged rows themselves rather than a count, 83,666 rows is a perfectly reasonable thing to `collect()`.

### The hard case

Only R has the function, and you need one row out per row in. There is no comfortable answer to this one, and it would be dishonest to manufacture one. Either the data crosses, or the function has to be made to run where the data is.

If your R function is genuinely irreducible and has to see every row, the honest options are to collect what you can, to narrow the input until it fits, or to send the function to the data. None of those is free.

Where this case usually turns out to be tractable is when the expensive part is not the crossing but the *repetition*: a small result crossed once, then simulated or resampled thousands of times. That has a good answer, and it is on [Run a Monte Carlo simulation](monte-carlo.qmd).

## Reaching a function `dbplyr` cannot translate, with `dplyr::sql()`

`dbplyr` not knowing a function is not the same as the server not having it. Those two get conflated constantly, and conflating them pushes work into R that never needed to leave the warehouse.

`grepl()` is the case you will meet first. `dbplyr` has no translation for it, so it goes into the SQL unchanged and fails on the server:

```{r}
#| label: fail
#| error: true
measures |>
  filter(grepl("mean", value_statistic)) |>
  summarise(n = n()) |>
  collect()
```

`UNRESOLVED_ROUTINE` is the signature. It says the query reached the server and the server did not recognise a function name, which is a narrower complaint than it looks: it is about one name, not about the shape of your query. Read that error as an invitation rather than a wall.

It does not, though, say *why* the name was not recognised, and there are three reasons it can appear. `dbplyr` had no translation and passed your R function through, which is this case. The function exists on Databricks but not on the compute you are connected to. Or you misspelled it. The message is the same for all three, differing only in the name it quotes, so read the name in the error before assuming which one you are in. [Which spatial functions you have](../ref/spatial-functions.qmd) covers the second, where the split is real and counter-intuitive.

The escape hatch is `dplyr::sql()`, which passes a fragment through untouched. You are writing Spark SQL inside it, so the dialect is theirs, not R's:

```{r}
#| label: escape
flow_measures <- measures |>
  filter(
    parameter == "flow",
    period == 86400L,
    sql("value_statistic LIKE '%/mean'")
  )
```

That is the same `flow_measures` the reduction above uses, and it still runs entirely on the server. The pattern match never became an R problem.

So when `dbplyr` refuses a function, the question to ask is whether Spark SQL has an equivalent, not whether you now have to collect the table. Usually it does, and `sql()` reaches it.

## Fitting a model on what came back

Once the data is in your session, there is nothing special about it. `per_station` is a 1,029-row tibble and everything you already know applies:

```{r}
#| label: stations
stations <- tbl(con, in_catalog(catalog, schema, "hydrology_stations"))

station_area <- stations |>
  select(station_id, catchment_area_km2) |>
  collect()

model_data <- per_station |>
  inner_join(station_area, by = "station_id") |>
  filter(catchment_area_km2 > 0, mean_flow > 0)

fit <- lm(log(mean_flow) ~ log(catchment_area_km2), data = model_data)
summary(fit)
```

That is the shrink-first case finishing. The second `collect()` brings back 9,536 rows of station metadata, the join happens in R because both sides are now small, and `lm()` never knew Databricks was involved.

## Where this stops

Not every model has a server-side counterpart, and the gap is worth naming precisely rather than as "some models".

Spark SQL has no ordinal-regression or cumulative-link primitive. If your response is an ordered factor, as ecological status classes are, there is nothing on the server to push the fit down to. The fit happens in R:

```{r}
#| label: polr
library(MASS)

catchments <- tbl(con, in_catalog(catalog, schema, "wfd_catchments"))

eco <- catchments |>
  filter(!is.na(ecological_class), ecological_class != "Not assessed") |>
  dplyr::select(water_body_id, ecological_class, water_body_area_m2) |>
  collect() |>
  mutate(
    class = factor(
      ecological_class,
      levels = c("Bad", "Poor", "Moderate", "Good", "High"),
      ordered = TRUE
    ),
    log_area = log(water_body_area_m2)
  )

polr(class ~ log_area, data = eco, Hess = TRUE)
```

This is fine, because 3,553 rows fit comfortably. It is the shrink-first case again: the selection and the filtering happened server-side, and only the columns the model needs crossed. The `dplyr::` prefix is there because attaching `MASS` masks `select()`, which is a nuisance you already know about and not a Databricks problem.

The reason to name the gap rather than gesture at it is that you can then check your own model the same way. Ask whether Spark SQL has a primitive for the fit you want. If it does not, the fit is an R step, and the question becomes how small you can make the data before it gets there.

## Next

If the repeats are the expensive part, see [Run a Monte Carlo simulation](monte-carlo.qmd).

---

This page rests on: a 32.5 million row table reduces by three orders of magnitude before crossing into R; `dplyr::sql()` reaches Spark SQL functions `dbplyr` cannot translate; `UNRESOLVED_ROUTINE` is raised identically for an untranslated function, a function absent from the connected compute, and a misspelling; Spark SQL has no ordinal-regression primitive.
