---
title: "Check that you get the same answer wherever it ran"
subtitle: "Confirming that a computation done on the server and the same one done in R agree, and by how much"
---

## The two sides agree, and here is how closely

Worker-computed polygon area matched the warehouse's own to a relative difference of 3.1e-12. So the answer to "can I trust work that ran somewhere else" is yes. That number is here to settle the question rather than to impress: agreement at the twelfth decimal place is agreement, and nothing downstream of it will notice the difference.

The honest limit belongs next to it. This is one comparison on one kind of computation. Until the same recipe has been run on a second, structurally different case, this page is teaching a technique on the strength of a single example, and you should read it that way. The recipe below is worth copying; the confidence that it generalises has not been earned yet.

## Why they can disagree at all

Four things make two correct machines return different numbers: floating point, aggregation order, a coordinate system you set in one place and not the other, and a spatial function that models the Earth differently on each side.

They are not equally likely and they are not equally serious, and telling them apart is most of what you need.

Floating point is the smallest. The warehouse and your R session are running different implementations of the same geometry library, compiled differently, on different hardware. Neither is wrong. The last few bits of a double simply do not have to match, and once you sum a few thousand of them they match slightly less.

Aggregation order is the next. When the server computes a sum, it computes it in partitions and adds the partial results together, and the partitioning is not guaranteed to be stable between runs. Floating-point addition is not associative, so a different order gives a different last bit. The same query run twice can differ from itself by this route, which is worth knowing before you conclude that the difference is between the two machines.

The coordinate system is the first one that produces a difference you can see, and it is a genuine mistake rather than a property of the platform. WKB carries no SRID, so nothing in the bytes tells `sf` which projection they are in. Set it wrong, or forget to set it, and you get a valid polygon with a completely wrong area. That difference is a factor, not a rounding artefact, and it is the first thing to check when a comparison fails.

The fourth is the largest and the only one where both machines are right. On longitude and latitude, `sf` measures on a sphere and returns metres, while the server discards the coordinate system and does flat arithmetic on the raw degrees. Compare those two and you get a difference of five orders of magnitude with no error on either side, because you asked two different questions and got two correct answers. [Join and transform polygons](polygons.qmd) has the detail and the fix, which is to project; here the thing to know is that a difference that large is more likely to be this than a bug.

So rank a discrepancy by its size before you go looking for a cause. A relative difference around 1e-12 is the first two causes and needs no action. Something visible in the first few significant figures is a CRS you set in one place and not the other. And a difference of orders of magnitude is usually not a mistake at all: it is the two sides measuring in different units, so check what each function actually returns before you start debugging.

Ranking by size has one blind spot, and it is worth knowing about before you trust the heuristic. The fourth cause has a quiet form that produces no magnitude at all: on unprojected coordinates, `sf` and the server also disagree about points sitting exactly on a boundary, and they agree everywhere else. What you see then is not a number that is wildly out but a count that is a few rows different, which the ranking above would send you looking for a bug in your own code. If you are comparing counts from a spatial join or a point-in-polygon test rather than comparing measurements, check the projection first and the tolerance second.

## The recipe

Compute the quantity on the server, compute it in R on data you can hold, compare the relative difference against a tolerance you have chosen, and write the tolerance down.

The catchment table makes this concrete because both sides can compute the same thing from the same bytes. Ask the warehouse for its own `ST_Area()` and for the geometry, then recompute the area in R.

Both calls are spelled `st_area()` below, and they are different functions. The one inside `transmute()` runs on the server, because everything above `collect()` does; the one inside `mutate()` is `sf`'s, because everything below it is. That is the whole point of the comparison, and it is worth seeing written out, since a spatial pipeline gives you no other clue which implementation answered.

Geometry is `BINARY`, so this needs the `brickster` connection rather than the ODBC one. [Connect your R session to the data](connect.qmd) covers why.

```{r}
#| label: mask
#| include: false
# This site is public, so no printed output may carry the catalog this page
# happens to run against. Only what is *shown* is masked; every query below
# runs against the real tables. See R/setup.R.
source(here::here("R/setup.R"))
wq_mask_output()
```

```{r}
#| label: setup
library(DBI)
library(dplyr)
library(dbplyr)
library(tibble)
library(cli)
library(sf)

catalog <- Sys.getenv("DATABRICKS_CATALOG")
schema <- Sys.getenv("DATABRICKS_SCHEMA")

acon <- dbConnect(
  brickster::DatabricksSQL(),
  warehouse_id = basename(Sys.getenv("DATABRICKS_PATH"))
)

both <- tbl(acon, in_catalog(catalog, schema, "wfd_catchments")) |>
  filter(!is.na(geom_wkb)) |>
  transmute(
    water_body_id,
    area_server = st_area(st_geomfromwkb(geom_wkb)),
    geom_wkb
  ) |>
  arrange(water_body_id) |>
  head(200) |>
  collect() |>
  mutate(
    area_r = st_as_sfc(structure(geom_wkb, class = "WKB"), EWKB = TRUE) |>
      st_set_crs(27700) |>
      st_area() |>
      as.numeric(),
    rel_diff = abs(area_r - area_server) / area_server
  ) |>
  select(area_server, area_r, rel_diff)

both
```

Two properties of that column are the point. The differences are around 1e-13, far below anything an area in square metres will be reported to. And they do not grow with the polygon: the largest relative difference in the run above is on a catchment of 257 square metres, the smallest on one of 250,000. That is what a floating-point artefact looks like. A CRS mistake would not behave this way, because it scales everything by roughly the same wrong factor.

Note what this comparison is not. The table also carries a `water_body_area_m2` column, recorded by whoever prepared the data, and your computed area will differ from that by a fraction of a percent rather than a fraction of a part per trillion. That is a useful check of a different thing: it tells you the bytes survived transport and the CRS is right. It cannot tell you whether two implementations of the same calculation agree, because the recorded column was not produced by either of them.

## Choosing a tolerance, and why "identical" is the wrong test

Exact equality fails on correct results. Write the test that way and it will report corruption that is not there, on data where nothing is wrong.

Run all three tests against the comparison above and only one of them passes:

```{r}
#| label: compare
tibble(
  test = c("identical()", "all(mine == theirs)",
           "max relative difference < 1e-9"),
  passes = c(
    identical(both$area_r, both$area_server),
    all(both$area_r == both$area_server),
    max(both$rel_diff) < 1e-9
  )
)
```

That is the whole argument. Both machines are correct, both answers are usable, and two of the three tests call it a failure. A check written on exact equality does not become more rigorous by being stricter; it becomes useless, because it fires every time and you stop reading it.

So the check is a relative difference against a tolerance. The tolerance is yours to choose and this page will not supply a constant, because the right value depends on what the number is for. A catchment area feeding a map wants something looser than a mass balance feeding a published figure. What matters is that you pick it deliberately, before you look at the result, and write it down next to the answer.

Wrap it so the failure says something useful:

```{r}
#| label: tolerance
agrees <- function(mine, theirs, tolerance) {
  rel <- abs(mine - theirs) / abs(theirs)
  worst <- max(rel, na.rm = TRUE)

  if (worst >= tolerance) {
    cli_abort(c(
      "Worst relative difference {signif(worst, 2)} exceeds {tolerance}.",
      i = "{sum(rel >= tolerance)} of {length(mine)} values fall outside it.",
      x = "Check the CRS on both sides before blaming rounding."
    ))
  }

  cli_alert_success(
    "{length(mine)} values agree, worst relative difference \\
     {signif(worst, 2)}, tolerance {tolerance}"
  )
  invisible(worst)
}

agrees(both$area_r, both$area_server, tolerance = 1e-9)
```

Reporting the worst difference on success, not just the pass, is what makes this worth running more than once. A run that passes at 1e-12 and later passes at 1e-4 has told you something changed, and a bare `TRUE` would have hidden it.

`all.equal()` is the base R version of the same idea and it is fine for an interactive look. It defaults to a tolerance of about 1.5e-8 and returns a message rather than `FALSE` on failure, so wrap it in `isTRUE()` if you branch on it. The reason to write your own is the tolerance: `all.equal()`'s default is a number somebody else chose, and the point of this section is that the number should be one you chose.

## What to check when the work ran in a worker rather than in your session

Everything above assumed the R side ran in your session. When it ran in a worker instead, there is one extra variable, and it is the worker's library.

Your session's packages and the worker's packages are separate installations. A different version of `sf`, or of the geometry library underneath it, can change a result legitimately: a bug fixed between versions, a different default, an algorithm changed. The number that comes back is then correct for the code that ran, and different from the number you would have got, and nothing about the difference looks like a version difference. It looks exactly like a bug.

So establish which versions ran before you try to explain a discrepancy. [Packages](../ref/packages.qmd) covers getting a worker's library into a state you can name. The order matters: a version mismatch you have not ruled out will send you debugging your own code for a difference that is not in it.

## Do not compare hashes

Hashing an R object is not portable across R minor versions, so a hash comparison between two machines can report a mismatch when the values are identical.

This is worth stating firmly because hashing is the obvious approach, and it fails in the most misleading possible way. A hash is designed so that any difference at all produces a completely different digest, which means a false mismatch is indistinguishable from real corruption. You get a wall of hex that agrees on nothing, and no way to tell whether the numbers underneath differ in the last bit or in the first.

It also cannot work here even when the R versions match, because of everything above. Two correct answers differ in the last bit, which is precisely what a hash is built to detect. A hash comparison would fail on the 200 catchments above, correctly by its own logic and uselessly by yours.

Compare content. If you need a compact summary rather than the full vectors, compare a few rounded aggregates: a sum, a range, a count, each with a stated tolerance. Those degrade gracefully, and they tell you how far apart two answers are rather than only that they are not the same.

## Recording the check alongside the result, so the paper can refer to it

Save the tolerance and the worst observed difference next to the answer, not in your head or your terminal scrollback. It is a row of data, so store it as one:

```{r}
#| label: record
checked <- tibble(
  quantity  = "catchment polygon area",
  n         = nrow(both),
  in_r      = "sf::st_area(), EPSG:27700",
  on_server = "ST_Area()",
  tolerance = 1e-9,
  worst     = max(both$rel_diff)
)

checked
```

Two reasons to bother. A reviewer asking how you know the server-side result is trustworthy gets a sentence and a number instead of an assurance. And the tolerance recorded before the run is the only thing that stops it drifting afterwards: a tolerance chosen once the answer is in front of you is not a check, it is a description.

Keep the row with the result it describes, in the same file or the same directory, and write it out at the same moment. [Get results out, and repeat them](results.qmd) covers where that file should go.

---

This page rests on: worker-computed polygon area matched the warehouse's own to a relative difference of 3.1e-12; hashing an R object is not portable across R minor versions; distance, length and area on the server's geometry type are planar while `sf` on unprojected coordinates is spherical, so the two disagree silently by orders of magnitude with neither being wrong; the same modelling difference makes predicates disagree on boundaries alone, which surfaces as a count a few rows out rather than as a difference in magnitude, so ranking a discrepancy by size does not find it; projecting the coordinates removes both, because `sf` uses s2 only for longitude and latitude.
