---
title: "Work out what the cluster you have can do"
subtitle: "Reading the compute you were given, and what you can and cannot change about it"
---

## Introduction

::: {.answer}
You can almost certainly start, stop and connect to the compute you have. You almost certainly cannot change how many machines it has, how much memory each gets, or which runtime it boots: those were decided when someone provisioned it.
:::

So the useful skill is not choosing a cluster. It is reading the one you have, and designing the work to fit it. Everything below is aimed at that, and any advice amounting to "use a bigger cluster" is beside the point.

## What you have

Most likely two things. A group cluster, shared with colleagues, that is probably already running because somebody else started it. And possibly a personal cluster that only you use, which you can start and stop yourself and which shuts down on its own after a period of inactivity.

The group cluster is warm and free to reach. The personal one is yours to control but costs minutes to wake. That difference drives most of the practical decisions on this page.

## Reading a cluster

Three lines, and you have the facts that matter:

```{r}
#| label: mask
#| include: false
# This site is public, so no printed output may carry the catalog this page
# happens to run against. Only what is *shown* is masked; every query below
# runs against the real tables. See R/setup.R.
source(here::here("R/setup.R"))
wq_mask_output()
```

```{r}
#| label: read-cluster
library(brickster)
library(tibble)

info <- db_cluster_get(cluster_id = Sys.getenv("DATABRICKS_CLUSTER_ID"))

tibble(
  workers = info$num_workers,
  runtime = info$spark_version,
  access  = info$data_security_mode
)
```

`num_workers = 0` means one machine. Not a small cluster, not a cluster with the workers switched off: one machine, running the driver, with nothing to distribute to. Whatever the workspace interface implies by showing it in a list of clusters, this is the number that settles it.

That single fact decides whether anything on [run a Monte Carlo simulation](monte-carlo.qmd) about distributing work applies to you at all. On a single-node cluster, distribution has nowhere to go, and reaching for it costs the serialisation without buying the parallelism.

One catch worth knowing before you trust the number. On a cluster configured to autoscale, `num_workers` is not `0`, it is absent entirely, because the count is a range rather than a value:

| Cluster | `num_workers` | `autoscale` | `ResourceClass` tag |
|---|---|---|---|
| Single-node | `0` | no | `SingleNode` |
| Autoscaling | `NULL` | yes | none |

So a test written as `num_workers == 0` returns `logical(0)` on an autoscaling cluster rather than `FALSE`, and an `if` on that errors. Check whether `autoscale` is present before reading `num_workers`, or read the `ResourceClass` tag, which says `SingleNode` on a genuinely single-machine cluster.

The same object carries the node type and the idle timeout, which are the other two things worth knowing about compute you did not choose.

## Access modes

A Standard access mode cluster refuses R **notebooks**. A client connection to the same cluster works.

That single word carries the whole claim, and it is worth being careful with because the unqualified version, the one you are likely to meet elsewhere, is simply false. "Standard access mode does not support R" reads as a wall. It is not one. If you believed it, you would abandon a cluster that would have served you and ask for a per-person cluster you never needed, which costs money and takes a conversation.

Your group cluster is quite likely Standard, which is exactly the case where this matters most. If someone has told you that you cannot use R on it, the thing to test is a client connection, not a notebook.

What is genuinely open: whether `spark_apply()` works under Standard access mode. This guide will not assert it either way, because nobody has established it. If you need distributed R specifically and your only cluster is Standard, that is a question to settle by trying it rather than a fact to look up.

## What costs time

| Situation | What it costs you |
|---|---|
| Starting a stopped personal cluster | Minutes. On the cluster used throughout this guide, terminated to running took between seven and eleven minutes across several starts |
| Connecting to an already-running cluster | Seconds. It is warm |
| Leaving a cluster running | It bills until it stops, and yours likely stops itself after the idle period in `autotermination_minutes` |
| Waking a SQL warehouse | Far less than an all-purpose cluster |

The first row is long enough to matter to how you plan an afternoon, and long enough that you should start the cluster before you need it rather than when you need it. The third tells you how long a coffee break can be before your next query pays that cost again, and it is on the object you already read above.

The last row is why a SQL warehouse is worth preferring for exploratory querying. Move to the cluster when you need something a warehouse cannot do, such as distributed R or a live Spark session.

## Which cluster answered

If you have more than one compute target configured, nothing in a result tells you which one produced it. A mistyped profile name does not fail, it silently falls back to the default, so the job runs on the wrong compute and returns something entirely plausible.

That matters most in the one situation where people keep two clusters: comparing a single-node run against a multi-node one. A clean run, a sensible number, and a confident wrong conclusion, because both halves of the comparison ran on the same machine.

So confirm which cluster answered rather than assuming. [Connect your R session to the data](connect.qmd) has the setup, including how to validate a profile name against the configuration file's own keys. This page is only here to say why you should care.

## What to ask for

If the shape of the cluster is genuinely wrong for your work, that is a request to whoever provisions compute, and it is worth making it once, completely.

Ask for the runtime version and the setup script together. They are one decision, not two. A cluster's setup script installs the packages your work needs, and it is compiled against the runtime. Change the runtime and leave the script attached and you get one of two outcomes: the packages are missing, or they recompile from source at every single start, turning a seven-minute wake into something much worse. Neither failure announces itself as a runtime mismatch.

So the request is "this runtime, with this setup script, tested together", not "please upgrade the runtime". [Setting up a cluster for geospatial R](../admin/geospatial-setup.qmd) is written for the person who will action it, and is meant to be forwarded.

One thing this page will not tell you: how work spread across worker machines scales. Distributing at all is one question and scaling is another, so if your plan depends on the second, measure it on your own cluster rather than taking a number from anywhere.
