---
title: "Connect your R session to the data"
subtitle: "Which connection to open, and why the choice matters before you write any code"
---

## Introduction

::: {.answer}
If your query touches a `BINARY` column, and geometry is stored as `BINARY`, open the connection with `brickster::DatabricksSQL()`. Otherwise `odbc::databricks()` is the default and is fine.
:::

That is the whole decision, and it is not a preference between two working options: the ODBC path cannot carry a `BINARY` value intact, so the moment geometry is in play the choice is made for you.

```{r}
#| label: mask
#| include: false
# This site is public, so no printed output may carry the catalog this page
# happens to run against. Only what is *shown* is masked; every query below
# runs against the real tables. See R/setup.R.
source(here::here("R/setup.R"))
source(here::here("R/facts.R"))
wq_mask_output()
```

```{r}
#| label: setup
library(DBI)
library(dplyr)
library(dbplyr)
library(tibble)
library(purrr)
library(glue)
library(cli)
library(sf)

catalog <- Sys.getenv("DATABRICKS_CATALOG")
schema <- Sys.getenv("DATABRICKS_SCHEMA")

# Ordinary queries through dbplyr.
con <- dbConnect(
  odbc::databricks(),
  httpPath = Sys.getenv("DATABRICKS_PATH"),
  DefaultStringColumnLength = 65535
)

# Anything reading a BINARY column.
acon <- dbConnect(
  brickster::DatabricksSQL(),
  warehouse_id = basename(Sys.getenv("DATABRICKS_PATH"))
)
```

Both point at the same warehouse, and both read your credentials from the environment rather than from anything you type. On Posit Workbench those variables are set for you when you sign in. If `DATABRICKS_HOST` is empty, you are not signed in, and no amount of connection debugging will fix that.

Two connections open at once is normal here. Use `con` for the work and reach for `acon` only where geometry crosses.

## Why: ODBC silently truncates `BINARY`

A `BINARY` value longer than 1,024 bytes arrives in R with `length mod 1024` bytes. A 1,025-byte polygon becomes 1 byte. A 5,670-byte one becomes 550. Nothing errors, nothing warns, and the fetch reports success. This is upstream issue `r-dbi/odbc#1024`.

The arithmetic is exact, which makes it easy to see once you look for it:

```{r}
#| label: truncation
tibble(asked = c(512, 1024, 1025, 2000, 5000)) |>
  mutate(returned = map_int(asked, \(n) {
    length(dbGetQuery(con, glue("SELECT unhex(repeat('AB', {n})) AS b"))$b[[1]])
  }))
```

Under 1,024 bytes everything is fine, which is the trap: a small geometry survives, so a test on simple shapes passes and the real data fails. On the catchment table used throughout this guide, 4,040 of `r wq_fact("catchments_rows")` catchment polygons exceed 1,024 bytes, and 91.5% of the stored geometry never reaches R.

`brickster::DatabricksSQL()` returns the same column byte-exact. Decode a polygon it fetched and the area agrees with the server's own `ST_Area()` to within a fraction of a square metre on a catchment of 25 million.

## The second trap, behind the first

The documented way round the `BINARY` bug is to encode the value as text on the server, with `base64()`, and decode it in R. Follow that alone and you walk into a different silent truncation.

Text columns are clipped at 1,023 characters by default. Encoded geometry runs well past that: on the same catchment table the longest encodes to over 200,000 characters. Measured across that table, the default connection discards about 87% of the encoded geometry while reporting success, and the failure surfaces as a decode error that reads like a corrupt payload rather than anything mentioning length.

This is not only about columns you read out of a table. Any long text value hits it, including one a server-side function just produced: asking for a computed geometry as text returned 1,023 characters of a value that was 114,609 long, and the truncated version still parsed. So it applies just as much to a spatial result you built in the query as to stored bytes, which is easy to miss because the workaround is usually explained as a way to read a `BINARY` column.

Both have to be worked around together. Encode server-side **and** raise the driver's string-length setting above the longest encoded value:

```{r}
#| label: fixedcon
#| eval: false
con <- dbConnect(
  odbc::databricks(),
  httpPath = Sys.getenv("DATABRICKS_PATH"),
  DefaultStringColumnLength = 65535
)
```

Use 65535 rather than a round number like 100000. Round values can be silently dropped, because they render in scientific notation on the way to the driver and arrive as something it does not read as a number. The setting is not a ceiling on the payload: with it in place a 206,338-character value arrives intact.

Both truncation bugs now have fix pull requests open against the driver package. Neither has merged, so the workarounds are needed today, and this section describes the situation until those fixes ship rather than a permanent property of the stack.

## Checking the connection

Both failures are silent, so the recipe is worth nothing without a check. Two checks, and they answer different questions.

Compare the length the server reports against the length that arrives. This catches truncation directly:

```{r}
#| label: octet
tbl(acon, in_catalog(catalog, schema, "wfd_catchments")) |>
  filter(!is.na(geom_wkb)) |>
  transmute(water_body_id, declared = octet_length(geom_wkb)) |>
  head(3) |>
  collect()
```

Then validate the decoded geometry against something recorded independently, rather than merely confirming it parses. The catchment table carries its own area column, computed before any of this transport happened, so comparing your computed area against it tests the whole round trip:

```{r}
#| label: areacheck
catchments <- tbl(acon, in_catalog(catalog, schema, "wfd_catchments")) |>
  select(water_body_id, water_body_area_m2, geom_wkb) |>
  head(5) |>
  collect() |>
  mutate(
    geometry = st_as_sfc(structure(geom_wkb, class = "WKB"), EWKB = TRUE),
    area_r = as.numeric(st_area(st_set_crs(geometry, 27700))),
    agreement = area_r / water_body_area_m2
  )

catchments |>
  select(water_body_area_m2, area_r, agreement)
```

The recomputed areas agree with the declared column to within a tenth of a percent. That is the answer you want: agreement at this level says the bytes survived and the CRS is right. It is not exact agreement, and it should not be, because the two numbers were computed by different software at different times. What matters is that the disagreement is small enough to be uninteresting and stays that way as you add rows.

A parse check on its own is weaker than it looks in both directions. On this data a truncated polygon usually fails to parse outright, so the loud error arrives before any check does. But that is a property of where the bytes happen to be cut, not a guarantee, and a geometry that parses can still be the wrong geometry: a silent CRS mismatch produces a perfectly valid polygon in the wrong place. The area comparison catches that too, which is why it is the check worth keeping.

## Finding your table

Names in Unity Catalog have three parts: catalog, schema, table. Read the first two from the environment rather than writing them into your script, so the same code runs against a different catalog without editing.

The three-part name is not bureaucracy. It is the unit that permission is granted on, which is why you can see a catalog listed and not see anything inside it. If a table you were told exists is missing, that is a grant, not a typo, and it is a question for whoever administers the workspace rather than something to debug in R.

## Configuring more than one compute target

If you work against a warehouse and a cluster, or two clusters, keep the identifiers in a configuration file rather than scattered through scripts. One trap makes this worth doing carefully: `config::get()` treats an unrecognised profile name as absent and falls back to the default profile without warning. A mistyped profile name does not fail. The job runs against whichever target the default names, returns results that look entirely ordinary, and nothing in the output says which compute answered.

So validate the name against the file's own keys instead of trusting it:

```{r}
#| label: profile
#| eval: false
compute_id <- function(profile, file = "compute.yml") {
  known <- names(yaml::yaml.load_file(file))

  if (!profile %in% known) {
    cli_abort(c(
      "Unknown profile {.val {profile}}.",
      i = "{.file {file}} defines: {.val {known}}.",
      x = "config::get() would have fallen back to
           {.val default} and given you the wrong compute."
    ))
  }

  config::get("cluster_id", config = profile, file = file)
}
```

Store the warehouse identifier on its own rather than as part of a longer connection path. Different clients want different forms of it: `odbc` wants the HTTP path, `brickster` wants the bare identifier. Storing the path means every `brickster` call takes it apart again with `basename()`, and the two drift out of step the first time one is edited and the other is not.

## What a connection is

Your connection is a client to remote compute, not a session on it. The R process is yours, running where you started it. `con` is a handle that sends SQL somewhere else and brings answers back.

That distinction decides what your code can do. An object in your R session is not visible to the warehouse, and a table in the warehouse is not in your memory until you ask for it. Everything else in this guide is a consequence of where that line falls and how much crosses it.

## Next

Whichever of these resembles the code you have written: [work with a table that will not fit in your session](big-table.qmd), [join and transform polygons](polygons.qmd), or [run a Monte Carlo simulation](monte-carlo.qmd).
