---
title: "Run a Monte Carlo simulation"
subtitle: "Simulating, bootstrapping, resampling or fitting many times over: once per site, per polygon, per resample or per replicate"
---

## Introduction

::: {.answer}
Your replicates are independent, so the question is not how much data moves but how many repeats run at once. Start on the cores you already have: one line turns `map()` into `future_map()`, and that is usually the whole answer.
:::

The expensive part of this work is not getting the data out. A large table shrinks server-side to a small result, which crosses cheaply, and only then does the real cost start: you do something to that result ten thousand times.

This page runs the repeats serially first, then across your own cores, then asks whether the cluster is worth it, which is often not. `spark_apply()` comes last, for the case where it is.

```{r}
#| label: facts
#| include: false
# One number in the closing summary is quoted from facts/measurements.yml
# rather than typed. Reads a local file and opens no connection.
source(here::here("R/facts.R"))
```

```{r}
#| label: setup
#| message: false
library(purrr)
library(tibble)
library(dplyr)
library(parallel)
```

## Is your problem this shape?

One independent unit per site, per polygon, per resample or per replicate.

The test: if replicate 200 does not need anything replicate 199 produced, the work is this shape, and everything below applies. Almost all simulation, bootstrapping and resampling is. If your replicates do depend on each other, in a Markov chain for instance, none of this applies and the page stops being useful to you.

Recognising the shape matters more than matching your problem to an example. You will know your own problem faster than you will find it in a list.

## Running them serially

Start here and keep it. The serial version is the one you can debug, and it is what every faster version has to agree with.

```{r}
#| label: serial
sim_one <- function(i) {
  set.seed(i)
  x <- rlnorm(200000, meanlog = 0, sdlog = 1)
  b <- map_dbl(1:60, \(k) mean(sample(x, length(x), replace = TRUE)))
  tibble(rep = i, boot_se = sd(b))
}

results <- map(1:24, sim_one) |> list_rbind()

head(results, 3)
```

## Running them on your own cores

One line changes:

```{r}
#| label: parallel
#| message: false
library(future)
library(furrr)

plan(multicore, workers = 2)

results <- future_map(1:24, sim_one, .options = furrr_options(seed = TRUE)) |>
  list_rbind()

head(results, 3)
```

That is the whole argument for trying your own cores first: `map()` becomes `future_map()`, you add a plan, and nothing else about the code moves. Given how little it costs to try, try it before you consider anything involving the cluster.

What it buys you depends on how many cores you actually have, which is the next section but one, and on how much work each replicate does. Time your own serial run and multiply: that arithmetic is more use than any figure this page could quote, because the machine building this page is not yours.

`furrr_options(seed = TRUE)` is not optional decoration. Without it the workers do not get independent, reproducible streams, and `furrr` will warn you about exactly that.

## Memory when forking

Forking to run the repeats in parallel does not raise peak memory. It lowers it.

```{r}
#| label: memory
peak_mb <- function(expr) {
  invisible(gc(reset = TRUE))
  force(expr)
  g <- gc()
  round(sum(g[, "max used"] * c(56, 8)) / 1e6, 1)
}

tibble(
  run = c("serial", "forked, 2 workers"),
  peak_mb = c(
    peak_mb(map(1:8, sim_one) |> list_rbind()),
    peak_mb(mclapply(1:8, sim_one, mc.cores = 2) |> list_rbind())
  )
)
```

Each worker holds only its own slice and returns a small summary, so the parent never has every intermediate in memory at once. This is the opposite of the big-table case, where pulling more rows into the session is exactly what hurts, and it is worth knowing because it removes a worry that would otherwise stop you parallelising.

## `multicore` versus `multisession`

`multisession` starts fresh R processes and copies what they need. `multicore` forks, so each child starts as a copy-on-write view of the parent and pays nothing for data it only reads.

That depends on forking. Under `multisession` each worker gets its own full copy, and the peak goes up rather than down.

`multisession` is the plan people reach for by default, partly because it is what works on Windows. You are on Linux here, where forking is available, so prefer `multicore` and know that the two are not interchangeable for this purpose.

## Choosing a worker count

Do not agonise over the number. The penalty for a wrong count is mild in both directions, and the fix is one number and a re-run.

Too few workers and you leave cores idle, which costs you some time. Too many and they contend, which costs you a little more. Neither corrupts anything. A reasonable guess, run it, adjust if it matters.

One caution specific to containers: `detectCores()` reports the machine's cores, not the share your container is allowed. The two do not always agree, and the direction that hurts is asking for more workers than your quota permits.

## Packages inside the worker

A forked worker inherits your session's loaded packages, so a local fork is fine. A worker on cluster compute is a different R process on a different machine, and it needs `sf` installed there.

Check before you build a pipeline on it, then see [Packages](../ref/packages.qmd). This depends on your administrator having built a library volume. If they have not, this section does not apply to you, and that is a conversation rather than something to solve in R.

## Bootstrapping and resampling

A bootstrap replicate does not need the previous one, so everything on this page applies unchanged.

## Tuning across resamples

`tune` already parallelises over resamples through the same `future` plan. The plan you set earlier applies, and there is no separate step to take.

So if you have set `plan(multicore, workers = 2)` and then call `tune_grid()`, the resamples spread across those workers. You do not pass anything to the tuning call, and you do not need a `tune`-specific option.

The corollary is worth holding on to: if you have not set a plan, tuning runs serially and nothing tells you so.

## Is the cluster worth it?

Often not. If your repeats finish on your own cores in an afternoon, that is the better route. Reaching for distribution costs setup time, adds a serialisation step, and introduces a second environment whose package library you do not control. None of that buys anything if the work already fits.

The arithmetic is usually enough to decide. Time one replicate, multiply by how many you need, divide by the cores you have. If the answer is hours, you are done thinking. If the answer is days, read on.

Sometimes the honest answer is that your laptop is the right machine. If the data fits and nobody dictates where the computation happens, local R is faster to set up and better for your science.

## Using `spark_apply()`

Last, and only once your own cores have been ruled out. That ordering is the recommendation, not a rhetorical device.

### On a single node

It was slower than forked workers, because it pays serialisation for no distribution benefit. On a single-node cluster there is nowhere for the work to go, so you buy the cost of shipping functions and data to a worker without buying any parallelism you did not already have.

### Across several machines

It does distribute: on a two-worker cluster the work ran on both, counted by distinct machine names. That answers whether it distributes at all and nothing about how it scales, so do not read it as a claim about forty machines or about what a shuffle costs.

### Count machines, not processes

Two machines can report the same process id, so a process count cannot tell a single-node cluster from a distributed one. If you want to know whether your work is actually spreading, have the worker return `Sys.info()[["nodename"]]` and count the distinct values.

### Naming the output columns

`spark_apply(columns = )` takes a Spark type string from a client connection and a plain character vector on the cluster's own R session. Neither form is rejected by the other path with a useful message, and the failure never names the argument: it surfaces as a complaint about a column, which reads as though your worker function returned the wrong thing.

Leaving the argument out works on both paths, which is what an example should do unless it needs to fix the types. [Write and run R against the cluster, without leaving your editor](interactive.qmd) shows both forms and the error the wrong one produces.

### The grouped path

`spark_apply(group_by = )` fails outright on `rpy2` 3.6.x. Where it did run, it used one R process where the default partitioning used four, and was slower on identical work for the same answer. The mechanism behind that is unexplained, so treat the grouped path as unavailable rather than as a slower alternative.
