---
title: "Join and transform polygons"
subtitle: "Real spatial work when the polygons start out in the table rather than in R"
---

## Introduction

::: {.answer}
Buffers, intersections, transforms, areas and predicates all run on the server, so a spatial question does not oblige you to pull geometry into your session first.
:::

This page covers where a spatial pipeline runs, how to project and join, the one modelling difference that makes `sf` and the server disagree, and how to bring geometry back when you do need it. Here are four thousand catchments unioned and measured, with one row crossing into R:

```{r}
#| label: mask
#| include: false
# This site is public, so no printed output may carry the catalog this page
# happens to run against. Only what is *shown* is masked; every query below
# runs against the real tables. See R/setup.R.
source(here::here("R/setup.R"))
source(here::here("R/facts.R"))
wq_mask_output()
```

```{r}
#| label: setup

suppressPackageStartupMessages({
  library(DBI)
  library(dplyr)
  library(dbplyr)
  library(tibble)
  library(cli)
  library(sf)
})

con <- dbConnect(
  odbc::databricks(),
  httpPath = Sys.getenv("DATABRICKS_PATH"),
  DefaultStringColumnLength = 65535
)

acon <- dbConnect(
  brickster::DatabricksSQL(),
  warehouse_id = basename(Sys.getenv("DATABRICKS_PATH"))
)
catalog <- Sys.getenv("DATABRICKS_CATALOG")
schema <- Sys.getenv("DATABRICKS_SCHEMA")
```

```{r}
#| label: catchments

catchments <- tbl(con, in_catalog(catalog, schema, "wfd_catchments"))

catchments |>
  filter(!is.na(geom_wkb)) |>
  mutate(g = st_setsrid(st_geomfromwkb(geom_wkb), 27700L)) |>
  summarise(
    n_catchments = n(),
    area_km2 = round(sum(st_area(g), na.rm = TRUE) / 1e6)
  ) |>
  collect()
```

Two things in that chunk are worth naming before you write your own.

Those `st_` calls are Databricks functions rather than `sf` ones, even where the name is identical. `dbplyr` does not recognise them, so it passes the name through to the server as text and never evaluates it in R. `st_area(g)` above `collect()` is the server's function; the same expression below `collect()` is `sf`'s. Attaching `sf` makes no difference either way. So `collect()` is not a stylistic choice in a spatial pipeline, it is the line that decides which implementation answers, and `show_query()` will tell you which side you are on: whatever it prints is the server's.

The two sides do not always agree, and one difference is silent. A name the server does not have fails loudly with `UNRESOLVED_ROUTINE`, which is the easy case. The one to plan around is a name that exists on both sides and means something different, and there is a single rule that avoids it: **work in a projected CRS**, which is the next section.

Two smaller things will interrupt you while porting an expression rather than mislead you. Return shapes differ: `st_contains()` gives one boolean per row against `sf`'s sparse index list, `st_distance()` a scalar per row against a matrix, and `st_buffer()` rejects `nQuadSegs`. And `st_makevalid()` has no server-side equivalent at all, which matters because repairing an invalid geometry is a routine step in `sf`: the server will tell you a polygon is invalid and will not fix it, so repair before loading or collect and repair in R. [Which spatial functions you have](../ref/spatial-functions.qmd) has the rest, including which functions your own compute is missing.

## Set the CRS, then project

The polygons are stored as WKB in a `BINARY` column, and WKB carries no SRID. Ask the server what coordinate system the stored bytes are in and it says zero, meaning it does not know:

```{r}
#| label: srid
catchments |>
  filter(!is.na(geom_wkb)) |>
  transmute(
    water_body_id,
    srid = st_srid(st_geomfromwkb(geom_wkb))
  ) |>
  head(3) |>
  collect()
```

The number has to come from you. For this data it is EPSG:27700, the British National Grid, because that is what the Environment Agency publishes. `st_setsrid()` asserts it server-side and `st_set_crs()` asserts it in R. Neither reprojects anything: they attach a label to coordinates that already had that meaning.

Skip the label and the two sides fail differently. The server refuses to transform geometry whose SRID it does not know, which is the helpful case. R computes anyway: decode a polygon, ask for its area without declaring anything, and you get numbers with no CRS, no warning and nothing wrong-looking about them:

```{r}
#| label: nocrs
rows <- tbl(acon, in_catalog(catalog, schema, "wfd_catchments")) |>
  filter(!is.na(geom_wkb)) |>
  select(geom_wkb) |>
  head(3) |>
  collect()

geom <- st_as_sfc(structure(rows$geom_wkb, class = "WKB"), EWKB = TRUE)

st_crs(geom)$epsg
as.numeric(st_area(geom))
```

Those areas happen to be right, because the stored coordinates are already metres. Feed the same CRS-less geometry to anything that needs to know where on Earth it is, though, and the silence ends: a distance in degrees, a join against a layer in another projection, a transform. The answer is wrong or absent and nothing says which.

So wrap `st_geomfromwkb()` in `st_setsrid()` every time, in the `mutate()` that first decodes the bytes, and never let an un-labelled geometry reach the verbs downstream of it.

Then project. On EPSG:27700 or any other metric projection, `sf` and Databricks agree. On longitude and latitude they do not, and nothing tells you. This data already satisfies that, which is why the areas above can be trusted; if your own layers are in lon/lat, `st_transform()` them first.

You already know that a distance across a sphere is not a distance across a flat map. The trap is that `sf` and Databricks pick opposite defaults, and the same line of code silently gets whichever default belongs to the side it ran on. `sf` treats lon/lat as points on a sphere and gives you metres. The server discards the coordinate system and does flat arithmetic on the raw numbers, so on lon/lat it hands you degrees. Both are correct answers to different questions.

Here is the size of the gap, both sides computed side by side, on two points five degrees apart:

```{r}
#| label: divergence
server <- dbGetQuery(con, "
  SELECT st_distance(a, b)       AS planar,
         st_distancesphere(a, b) AS sphere
  FROM (SELECT st_geomfromtext('POINT(0 0)', 4326) AS a,
               st_geomfromtext('POINT(3 4)', 4326) AS b)
")

in_r <- st_distance(
  st_sfc(st_point(c(0, 0)), crs = 4326),
  st_sfc(st_point(c(3, 4)), crs = 4326)
)

tibble(
  asked = c("server, st_distance", "server, st_distancesphere", "sf, in R"),
  answer = c(server$planar, server$sphere, as.numeric(in_r))
)
```

Five against half a million: degrees against metres, each computed correctly. On an area it is worse, because square degrees are not a unit anyone would notice in a table of catchment sizes.

If you do want a distance across a sphere from lon/lat, ask for it by name: `st_distancesphere()` and `st_distancespheroid()`, as the comparison above shows. Sphere or ellipsoid is the ordinary choice you already make in `sf`.

### Spatial joins

The natural conclusion from the numbers above is that measurements are risky and true-or-false tests are safe. That is wrong, and it is the more expensive half.

On lon/lat, `st_contains()` and its relatives agree with `sf` in a polygon's interior and disagree **on its boundary**. In `sf` a polygon edge between two lon/lat corners is a great-circle arc, so it bows away from the straight line the server draws between the same two points. A point in that sliver is inside for one and outside for the other.

Which is exactly where a catchment join lands its difficult records. Catchments share their edges, coastlines are boundaries, and an outfall on a river is on a catchment boundary rather than comfortably inside one. So a point-in-polygon count on lon/lat comes back a handful of rows different from what `sf` would have said, every number in it is plausible, and the rows that moved are the ones you would have wanted to inspect.

Project both layers and the disagreement goes away, because `sf` only reaches for spherical geometry when the data is unprojected: on a projected CRS it works on a plane, which is what the server was doing all along.

### What does not fix it

Two things look like fixes and are not.

`sf_use_s2(FALSE)` is the first thing an experienced `sf` user reaches for, and it is a trap. It does not switch `sf` to a flat model, it switches it to an ellipsoidal one. So the boundary disagreement clears up and it looks as though it worked, while the measurements stay wrong by five orders of magnitude. You have cured the symptom you could have caught and kept the one you could not.

Tagging the SRID is not the same as respecting it. Keep wrapping `st_geomfromwkb()` in `st_setsrid()`, but `st_distance()` ignores the SRID you attached. A column that carries a CRS has not been measured in it.

Nothing will warn you about either, because whether a call is right depends on the CRS of the data and the SQL is written before anything asks the server what it is holding.

## Count the points inside the polygons

This is the shape most analyses need: two layers, a predicate, and a count coming back. Counting storm overflows inside catchments, grouped by the catchment's ecological class, returns five rows and leaves both layers where they are.

The overflows are longitude and latitude, EPSG:4326, and the catchments are EPSG:27700, so one of them has to move first. `st_transform()` runs server-side, so that reprojection happens once inside the query rather than in R on every pass:

```{r}
#| label: transform
tbl(con, in_catalog(catalog, schema, "storm_overflows")) |>
  filter(!is.na(longitude)) |>
  mutate(p = st_transform(st_point(longitude, latitude, 4326L), 27700L)) |>
  transmute(
    overflow_id,
    easting = round(st_x(p), 1),
    northing = round(st_y(p), 1)
  ) |>
  head(3) |>
  collect()
```

`st_point()` takes the SRID as its third argument, which is what makes the transform legal.

That transform is doing two jobs, and the second is the one to notice: it moves the overflows out of longitude and latitude before any predicate touches them, which is what makes the count below agree with the answer `sf` would give. It is not tidiness, it is the correctness step from the section above.

A spatial predicate cannot be a `join_by()` condition, which is the mechanical blocker you will hit first. `join_by()` accepts only equality and a few range helpers. Pair the two tables with `cross_join()` and filter on the predicate instead. That reads as though every catchment is about to be compared with every overflow, and the planner does not execute it that way: the filter becomes the join condition, exactly as it would have been in SQL.

```{r}
#| label: join
overflows <- tbl(con, in_catalog(catalog, schema, "storm_overflows")) |>
  mutate(p = st_transform(st_point(longitude, latitude, 4326L), 27700L))

catchments |>
  filter(!is.na(ecological_class)) |>
  mutate(g = st_setsrid(st_geomfromwkb(geom_wkb), 27700L)) |>
  cross_join(overflows) |>
  filter(st_contains(g, p)) |>
  count(ecological_class, name = "n_overflows") |>
  arrange(desc(n_overflows)) |>
  collect()
```

Both SRIDs are asserted inside the join condition, and that is not decoration: `st_contains()` refuses two arguments with different SRIDs rather than assuming they match.

Reading the overflow data here is subject to its own licence, which is CC BY 4.0 and not the Open Government Licence that covers the rest.

The worked example does this same join the other way, in `sf` after collecting both layers, and then checks the two answers against each other. See [The spatial join](../example/spatial.qmd) if you want the comparison rather than the recipe.

## Aggregating server-side

The shape to look for in your own analysis: the geometry stays where it is, the spatial work happens there, and a small table comes back.

Group by an ordinary column while aggregating the geometry. Here that is ecological class, six values, with the catchments in each class unioned and the union's area measured:

```{r}
#| label: aggregate
catchments |>
  filter(!is.na(geom_wkb), !is.na(ecological_class)) |>
  mutate(g = st_setsrid(st_geomfromwkb(geom_wkb), 27700L)) |>
  group_by(ecological_class) |>
  summarise(
    n = n(),
    area_km2 = round(st_area(st_union_agg(g)) / 1e6)
  ) |>
  arrange(desc(area_km2)) |>
  collect()
```

Six rows crossed. The 27 MB of polygons behind them did not.

One narrow restriction is worth knowing, because it is easy to meet and easy to work around. Grouping *by* a geometry column fails, since the type has no ordering. Grouping by an ordinary key while aggregating geometry works, as above, and so does grouping by something derived from the geometry such as a geohash.

## Bring the polygons into R

Collecting the geometry is a legitimate answer, not a defeat. If it fits in your session and you have an `sf` pipeline already written, this is the right call: you get the whole of `sf` rather than the subset the server exposes, and you are working in an idiom you already know.

The connection decides how. Geometry cannot travel over the ODBC path, because `odbc` silently returns only part of a `BINARY` value, so this route needs `brickster::DatabricksSQL()`. See [Connect your R session to the data](connect.qmd) for why, and for the check that the bytes arrived whole.

```{r}
#| label: collect-sf
catchment_sf <- tbl(acon, in_catalog(catalog, schema, "wfd_catchments")) |>
  filter(!is.na(geom_wkb)) |>
  select(water_body_id, ecological_class, water_body_area_m2, geom_wkb) |>
  collect() |>
  mutate(
    geometry = st_as_sfc(structure(geom_wkb, class = "WKB"),
                         EWKB = TRUE) |>
      st_set_crs(27700)
  ) |>
  select(-geom_wkb) |>
  st_as_sf()

catchment_sf |> select(water_body_id, ecological_class)
```

All `r wq_fact("catchments_rows")` polygons, 27.2 MB in the session, and the projection named in the header rather than left blank. That is what a good outcome looks like: `Projected CRS: OSGB36 / British National Grid` is the line to check before going any further, because the alternative is a blank where it should be and no other symptom.

Twenty-seven megabytes is small, and the honest reading of that is that this dataset never needed a decision. If your polygons are of that order and nobody is dictating where the analysis lives, your laptop will do this faster than any of the above and you should not think about it again. The server-side route earns its complexity when the geometry is large and the answer is small, not when both are small.

## Next

Gridded data does not work like this. The `ST_` functions are vector functions with no raster equivalent, so [work with rasters](terra.qmd) inverts everything on this page: there is no server-side path, and the work comes to you. And if you want to confirm the server-side answer matches R's, see [Check that you get the same answer wherever it ran](check-the-answer.qmd).
