---
title: "Join and transform polygons"
subtitle: "Real spatial work when the polygons start out in the table rather than in R"
---

## The spatial functions are on both sides of the line

The server has getting on for a hundred built-in `ST_` functions, so needing a spatial operation is not by itself a reason to bring data back. Buffers, intersections, transforms, areas and predicates all run where the data is.

That matters because the obvious chain of reasoning is wrong. Spatial work means `sf`, `sf` means R, R means collecting everything first: each step follows from the last and the conclusion is that geometry always crosses. It does not. You can put a spatial question to the server and get back a number.

Here is the shape, on the catchment polygons used throughout this guide. Four thousand polygons are read, unioned and measured, and one row comes back. Two connections are open because this page needs both: `con` runs the spatial work, and `acon` is the one that can carry raw geometry back, for the reason [Connect your R session to the data](connect.qmd) sets out.

The `st_` functions below are Databricks functions, not `sf` ones, even though several share a name. They are written inside an ordinary `dplyr` pipeline: `dbplyr` passes a function it does not recognise straight through to the server, so a spatial expression needs no translation to push down.

Worth being clear about what that means, because it is the one thing on this page that can mislead you. Inside a `dplyr` verb on a remote table, `st_area(g)` is not a call to `sf::st_area()`. Nothing in R evaluates it: `dbplyr` captures the name and sends it as text, and the server resolves it. Attaching `sf` changes nothing, and detaching it changes nothing. The same expression below `collect()` is the `sf` function, and above it is the server's.

A name the server does not have fails loudly, with `UNRESOLVED_ROUTINE`, whether it is a typo or a real `sf` function with no server equivalent. That is the safe case, and it is not the one to plan around.

The case to plan around is a name that exists on both sides and does not mean the same thing, because that runs cleanly and answers wrongly. There is one thing to remember about it, and [the next section](#project-before-you-measure-or-the-server-will-answer-in-degrees) is that thing: **work in a projected CRS, and the two sides agree.** On longitude and latitude they diverge, in measurements and in spatial joins alike, and nothing errors either way.

A third difference is harmless by comparison, and worth knowing only because it will interrupt you while porting an expression: `st_contains()` returns one boolean per row on the server against a sparse index list in `sf`, `st_distance()` a scalar per row against a matrix, and `st_buffer()` takes two arguments and rejects `sf`'s `nQuadSegs`. You will see each of those immediately. [Which spatial functions you have](../ref/spatial-functions.qmd#where-the-semantics-differ-from-sf) tabulates them.

So the position of `collect()` is not a stylistic detail in a spatial pipeline: it decides which implementation of a shared name you get. When you are unsure which side you are on, `show_query()` settles it, and anything it shows you is the server's.

```{r}
#| label: mask
#| include: false
# This site is public, so no printed output may carry the catalog this page
# happens to run against. Only what is *shown* is masked; every query below
# runs against the real tables. See R/setup.R.
source(here::here("R/setup.R"))
source(here::here("R/facts.R"))
wq_mask_output()
```

```{r}
#| label: setup
library(DBI)
library(dplyr)
library(dbplyr)
library(tibble)
library(cli)
library(sf)

catalog <- Sys.getenv("DATABRICKS_CATALOG")
schema <- Sys.getenv("DATABRICKS_SCHEMA")

con <- dbConnect(
  odbc::databricks(),
  httpPath = Sys.getenv("DATABRICKS_PATH"),
  DefaultStringColumnLength = 65535
)

acon <- dbConnect(
  brickster::DatabricksSQL(),
  warehouse_id = basename(Sys.getenv("DATABRICKS_PATH"))
)

catchments <- tbl(con, in_catalog(catalog, schema, "wfd_catchments"))

catchments |>
  filter(!is.na(geom_wkb)) |>
  mutate(g = st_setsrid(st_geomfromwkb(geom_wkb), 27700L)) |>
  summarise(
    n_catchments = n(),
    area_km2 = round(sum(st_area(g), na.rm = TRUE) / 1e6)
  ) |>
  collect()
```

How many exactly depends on the compute, and not in the direction you would guess. There were `r wq_fact("spatial_functions_warehouse")` on a Pro warehouse running serverless when this was measured, and `r wq_fact("spatial_functions_cluster")` on an all-purpose cluster. `st_makepoint()`, `st_makeenvelope()`, `st_pointonsurface()` and `st_collect()` are the four missing from the cluster. So the larger, more expensive compute has the smaller spatial vocabulary. If you need a point from two numbers on a cluster, `st_point()` works on both and does the same job. The older Classic warehouse type is a third case, where the functions are documented as unavailable, per function rather than anywhere you would find it by browsing.

`r wq_fact("spatial_functions_warehouse", "words")` is a lot, and it is still not all of `sf`. The absences worth knowing before you plan around them are `st_makevalid()`, `st_relate()`, `st_voronoi()`, `st_snap()`, `st_segmentize()`, `st_sample()`, `st_polygonize()` and `st_shortestline()`. `st_makevalid()` is the one likely to stop you, because repairing an invalid geometry is a routine step in `sf` and there is no server-side equivalent: the server will tell you a polygon is invalid with `st_isvalid()` and will not fix it. Repair before loading, or collect and repair in R.

That leaves you one ambiguous error to interpret. A function that does not exist on your compute fails with the same message as a name you misspelled, differing only in the name it quotes, so it cannot tell you which mistake you made. Check the spelling and the compute together, because the platform will not separate them for you. A wrong number of arguments is the easy case, and does produce a distinguishable error naming how many the function wanted.

## Set the CRS yourself

The polygons are stored as WKB in a `BINARY` column, and WKB carries no SRID. Ask the server what coordinate system the stored bytes are in and it says zero, meaning it does not know:

```{r}
#| label: srid
catchments |>
  filter(!is.na(geom_wkb)) |>
  transmute(
    water_body_id,
    srid = st_srid(st_geomfromwkb(geom_wkb))
  ) |>
  head(3) |>
  collect()
```

The number has to come from you. For this data it is EPSG:27700, the British National Grid, because that is what the Environment Agency publishes. `st_setsrid()` asserts it server-side and `st_set_crs()` asserts it in R. Neither reprojects anything: they attach a label to coordinates that already had that meaning.

Skip it and the two sides fail differently, which is the part worth knowing before you pick where to work. R computes anyway. Decode a polygon, ask for its area without declaring anything, and you get numbers with no CRS, no warning and nothing wrong-looking about them:

```{r}
#| label: nocrs
rows <- tbl(acon, in_catalog(catalog, schema, "wfd_catchments")) |>
  filter(!is.na(geom_wkb)) |>
  select(geom_wkb) |>
  head(3) |>
  collect()

geom <- st_as_sfc(structure(rows$geom_wkb, class = "WKB"), EWKB = TRUE)

st_crs(geom)$epsg
as.numeric(st_area(geom))
```

Those areas happen to be right, because the stored coordinates are already metres. Feed the same CRS-less geometry to anything that needs to know where on Earth it is, though, and the silence ends: a distance in degrees, a join against a layer in another projection, a transform. The answer is wrong or absent and nothing says which.

The server is stricter, and its strictness is a gift here. It refuses rather than guessing. Ask it to transform geometry whose SRID it does not know, and nothing happens at all:

```{r}
#| label: nosrid
#| error: true
catchments |>
  filter(!is.na(geom_wkb)) |>
  transmute(g = st_transform(st_geomfromwkb(geom_wkb), 27700L)) |>
  head(1) |>
  collect()
```

So the habit to build is to wrap `st_geomfromwkb()` in `st_setsrid()` every time, in the `mutate()` that first decodes the bytes, and never let an un-labelled geometry reach the verbs downstream of it.

Necessary, though, is not sufficient, and this is the point at which it would be reasonable to think the CRS problem is solved. It is not: labelling the coordinates does not make the server measure them the way `sf` would. The next section is the part that bites.

## Project before you measure, or the server will answer in degrees

**Work in a projected CRS whenever a spatial expression runs on the server.** On EPSG:27700 or any other metric projection, `sf` and Databricks agree. On longitude and latitude they do not, and nothing tells you.

That is the whole rule, and this data already satisfies it: the catchments are stored in EPSG:27700, which is why the areas above can be trusted. If your own layers are in lon/lat, `st_transform()` them first, and the rest of this section is why.

Setting the CRS is necessary and not sufficient, which is the part that catches people. Labelling the coordinates does not change how they are measured.

You already know that a distance across a sphere is not a distance across a flat map. The trap is that `sf` and Databricks pick opposite defaults, and the same line of code silently gets whichever default belongs to the side it ran on. `sf` treats lon/lat as points on a sphere and gives you metres. The server discards the coordinate system and does flat arithmetic on the raw numbers, so on lon/lat it hands you degrees. Both are correct answers to different questions.

Here is the size of the gap, both sides computed side by side, on two points five degrees apart:

```{r}
#| label: divergence
server <- dbGetQuery(con, "
  SELECT st_distance(a, b)       AS planar,
         st_distancesphere(a, b) AS sphere
  FROM (SELECT st_geomfromtext('POINT(0 0)', 4326) AS a,
               st_geomfromtext('POINT(3 4)', 4326) AS b)
")

in_r <- st_distance(
  st_sfc(st_point(c(0, 0)), crs = 4326),
  st_sfc(st_point(c(3, 4)), crs = 4326)
)

tibble(
  asked = c("server, st_distance", "server, st_distancesphere", "sf, in R"),
  answer = c(server$planar, server$sphere, as.numeric(in_r))
)
```

Five against half a million: degrees against metres, each computed correctly. On an area it is worse, because square degrees are not a unit anyone would notice in a table of catchment sizes.

### Spatial joins are affected too, and that is the one that will catch you

The natural conclusion from the numbers above is that measurements are risky and true-or-false tests are safe. That is wrong, and it is the more expensive half.

On lon/lat, `st_contains()` and its relatives agree with `sf` in a polygon's interior and disagree **on its boundary**. In `sf` a polygon edge between two lon/lat corners is a great-circle arc, so it bows away from the straight line the server draws between the same two points. A point in that sliver is inside for one and outside for the other.

Which is exactly where a catchment join lands its difficult records. Catchments share their edges, coastlines are boundaries, and an outfall on a river is on a catchment boundary rather than comfortably inside one. So a point-in-polygon count on lon/lat comes back a handful of rows different from what `sf` would have said, every number in it is plausible, and the rows that moved are the ones you would have wanted to inspect.

Project both layers and the disagreement goes away, because `sf` only reaches for spherical geometry when the data is unprojected: on a projected CRS it works on a plane, which is what the server was doing all along.

### Two things that look like the fix and are not

**`sf_use_s2(FALSE)` is a trap**, and it is the first thing an experienced `sf` user reaches for. It does not switch `sf` to a flat model, it switches it to an ellipsoidal one. So the boundary disagreement clears up and it looks as though it worked, while the measurements stay wrong by five orders of magnitude. You have cured the symptom you could have caught and kept the one you could not. Leave s2 alone and project instead.

**Tagging the SRID is not the same as respecting it.** Keep wrapping `st_geomfromwkb()` in `st_setsrid()`, for the reasons in the section above, but `st_distance()` ignores the SRID you attached. A column that carries a CRS has not been measured in it.

Nothing will warn you about either. Whether a spatial call is right depends on the CRS of the data, and the SQL is written before anything asks the server what it is holding, so no check can catch this for you. Attaching `sf` makes no difference, and neither does the connection you opened.

If you do want a distance across a sphere from lon/lat, ask for it by name: `st_distancesphere()` and `st_distancespheroid()`, as the comparison above shows. Sphere or ellipsoid is the ordinary choice you already make in `sf`.

The geography type is a different route rather than a cast. `st_distance()` refuses a geography argument outright, so you cannot repair a flat answer by converting the input: use the two functions just named, or work in geography throughout with `st_length()` and `st_area()`, which do accept it. It is also still Public Preview where the geometry type is generally available.

Mixing two coordinate systems is the kind case by comparison: that fails loudly, as [the section above](#set-the-crs-yourself) shows. This one does not fail at all.

## Transforming once, server-side, instead of per iteration

`st_transform()` works server-side, so a reprojection can happen once, in the query, rather than in R on every pass of an analysis. The storm overflow locations are longitude and latitude, EPSG:4326, and the catchments are EPSG:27700, so something has to move before the two layers can meet:

```{r}
#| label: transform
tbl(con, in_catalog(catalog, schema, "storm_overflows")) |>
  filter(!is.na(longitude)) |>
  mutate(p = st_transform(st_point(longitude, latitude, 4326L), 27700L)) |>
  transmute(
    overflow_id,
    easting = round(st_x(p), 1),
    northing = round(st_y(p), 1)
  ) |>
  head(3) |>
  collect()
```

Note that `st_point()` takes the SRID as its third argument, which is what makes the transform legal. The same rule as above: say what the coordinates are before asking anything to move them.

This transform is doing two jobs, and the second is easy to miss. The layers have to share a coordinate system before they can meet, which is the obvious one. It also moves the overflows out of longitude and latitude before any predicate touches them, which is what makes the join at the foot of this page agree with the answer `sf` would give. Reprojecting to suit the join is the same act as reprojecting for correctness, so this pattern is worth keeping even when both layers already match.

## Aggregating server-side when only a scalar needs to cross

This is the best case on the site, and the shape to look for in your own analysis: the geometry stays where it is, the spatial work happens there, and a small table comes back.

Group by an ordinary column while aggregating the geometry. Here that is ecological class, six values, with the catchments in each class unioned and the union's area measured:

```{r}
#| label: aggregate
catchments |>
  filter(!is.na(geom_wkb), !is.na(ecological_class)) |>
  mutate(g = st_setsrid(st_geomfromwkb(geom_wkb), 27700L)) |>
  group_by(ecological_class) |>
  summarise(
    n = n(),
    area_km2 = round(st_area(st_union_agg(g)) / 1e6)
  ) |>
  arrange(desc(area_km2)) |>
  collect()
```

Six rows crossed. The 27 MB of polygons behind them did not.

You may have read that geometry cannot take part in grouped queries. The restriction is much narrower than that reads. Grouping by an ordinary key while aggregating geometry works, as above, and so does grouping by something derived from the geometry such as a geohash. What fails is grouping by a geometry column itself, because the type has no ordering. The error says exactly that and names the column:

```
✖  (80) Syntax or semantic analysis error thrown in server while
  executing query. Error message from server:
  org.apache.spark.sql.catalyst.ExtendedAnalysisException: The
  expression "g" cannot be used as a grouping e
```

So the pattern is not blocked, only one spelling of it is, and the one you want is the spelling that works.

## The full `sf` pipeline below the line, when that is what you want

Bringing the polygons back is a legitimate answer, not a defeat. If the geometry fits in your session and you have an `sf` pipeline already written, collecting it is the right call: you get the whole of `sf` rather than the subset the server exposes, and you are working in an idiom you already know.

The connection caveat decides how. Geometry cannot travel over the ODBC path, because `odbc` silently returns only part of a `BINARY` value, so this route needs `brickster::DatabricksSQL()`. See [Connect your R session to the data](connect.qmd) for why, and for the check that the bytes arrived whole.

```{r}
#| label: collect-sf
catchment_sf <- tbl(acon, in_catalog(catalog, schema, "wfd_catchments")) |>
  filter(!is.na(geom_wkb)) |>
  select(water_body_id, ecological_class, water_body_area_m2, geom_wkb) |>
  collect() |>
  mutate(
    geometry = st_as_sfc(structure(geom_wkb, class = "WKB"),
                         EWKB = TRUE) |>
      st_set_crs(27700)
  ) |>
  select(-geom_wkb) |>
  st_as_sf()

catchment_sf |> select(water_body_id, ecological_class)
```

All `r wq_fact("catchments_rows")` polygons, 27.2 MB in the session, and the projection named in the header rather than left blank. That is what a good outcome looks like: `Projected CRS: OSGB36 / British National Grid` is the line to check before going any further, because the alternative is a blank where it should be and no other symptom.

Twenty-seven megabytes is small, and the honest reading of that is that this dataset never needed a decision. If your polygons are of that order and nobody is dictating where the analysis lives, your laptop will do this faster than any of the above and you should not think about it again. The server-side route earns its complexity when the geometry is large and the answer is small, not when both are small.

## Points work the same way

Nothing above is specific to polygons. Points get the same treatment, and the predicate join runs server-side too. Counting storm overflows inside catchments, grouped by the catchment's ecological class, takes both layers, transforms one of them and returns five rows.

One wrinkle: `join_by()` accepts only equality and a few range helpers, so a spatial predicate cannot be written as the join condition. Pair the two tables with `cross_join()` and filter on the predicate instead. That reads as though every catchment is about to be compared with every overflow, and the query planner does not execute it that way: the filter becomes the join condition, exactly as it would have been in SQL.

```{r}
#| label: join
overflows <- tbl(con, in_catalog(catalog, schema, "storm_overflows")) |>
  mutate(p = st_transform(st_point(longitude, latitude, 4326L), 27700L))

catchments |>
  filter(!is.na(ecological_class)) |>
  mutate(g = st_setsrid(st_geomfromwkb(geom_wkb), 27700L)) |>
  cross_join(overflows) |>
  filter(st_contains(g, p)) |>
  count(ecological_class, name = "n_overflows") |>
  arrange(desc(n_overflows)) |>
  collect()
```

Both SRIDs are asserted inside the join condition, and that is not decoration: `st_contains()` refuses two arguments with different SRIDs rather than assuming they match.

Both layers are also in EPSG:27700 by the time the predicate sees them, and that is what makes this count trustworthy rather than approximately right. Had the overflows still been in longitude and latitude, this join would have run without complaint and disagreed with `sf` on every point sitting on a catchment boundary, which on a layer of catchments that share their edges is not a rare case. The transform is load-bearing, not tidiness.

Reading the overflow data here is subject to its own licence, which is CC BY 4.0 and not the Open Government Licence that covers the rest.

The worked example does this same join the other way, in `sf` after collecting both layers, and then checks the two answers against each other. See [The spatial join](../example/spatial.qmd) if you want the comparison rather than the recipe.

## Next

Gridded data does not work like this. The `ST_` functions are vector functions with no raster equivalent, so [work with rasters](terra.qmd) inverts everything on this page: there is no server-side path, and the work comes to you. And if you want to confirm the server-side answer matches R's, see [Check that you get the same answer wherever it ran](check-the-answer.qmd).

---

This page rests on: the platform reports `r wq_fact("spatial_functions_warehouse")` distinct built-in `ST_` functions on a Pro warehouse running serverless, and `r wq_fact("spatial_functions_cluster")` on an all-purpose cluster, where `st_makepoint`, `st_makeenvelope`, `st_pointonsurface` and `st_collect` are absent; `st_makevalid`, `st_relate`, `st_voronoi` and about twenty other `sf` staples are absent from both; an unknown-function error differs only in the name it quotes, whether the function exists on other compute or was never a function at all; `st_distance` rejects a geography argument, so geography is not a cast away from a flat answer, while `st_distancesphere` and `st_distancespheroid` return the sphere's and the ellipsoid's answers in metres where the planar call returns the separation in degrees; distance, length and area on the geometry type are planar in the coordinate system's own units, where `sf` on unprojected coordinates returns spherical metres, and the two differ silently by five orders of magnitude; the cause is one modelling difference, `sf` using s2 spherical geometry for longitude and latitude since version 1.0 while the server discards the coordinate system and works planar, so the same difference also makes predicates diverge exactly on boundaries, where a point on the interior of a polygon edge is TRUE to `st_intersects`, `st_covers` and `st_touches` on the server and FALSE in `sf`, while interior and exterior points agree, meaning "measures are unsafe and predicates are safe" is false; projecting the coordinates removes both divergences because `sf` engages s2 only for longitude and latitude, and every case tested then agreed exactly; neither attaching `sf`, nor tagging the SRID, nor changing connection path between `odbc`, `brickster` and `sparklyr` avoids it, and no automated guard can, because correctness depends on the SRID and `dbplyr` builds the SQL before consulting the server; `sf_use_s2(FALSE)` is a trap rather than a fix, switching to an ellipsoidal rather than a planar model, so it repairs the predicates while leaving the measures wrong by five orders of magnitude; twenty-two `sf` export names exist on both sides while the other ninety-four fail loudly, counted once against one runtime and one `sf` version; `st_transform` works server-side between EPSG:4326 and EPSG:27700; geometry is stored as WKB with no embedded SRID; polygon area aggregates server-side with only a scalar crossing; `dbplyr` sends an untranslated function name to the server verbatim, so a spatial expression pushes down from an ordinary `dplyr` pipeline and is never evaluated by `sf`, whether or not `sf` is attached; a spatial predicate cannot be a `join_by()` condition, and `cross_join()` followed by a filter on the predicate is planned as the same join.
