---
title: "Simulate on the cores you have"
subtitle: "Simulating, bootstrapping, resampling or fitting many times over, on the cores in your own session"
---

## Introduction

::: {.answer}
Your replicates are independent, so the question is not how much data moves but how many repeats run at once. Start on the cores you already have: one line turns `map()` into `future_map()`, and that is usually the whole answer.
:::

The expensive part of this work is not getting the data out. A large table shrinks server-side to a small result, which crosses cheaply, and only then does the real cost start: you do something to that result ten thousand times.

This page runs the repeats serially first, then across the cores in your own session, and ends at the arithmetic that tells you whether you need anything more. Most of the time you do not. Where you do, [Run the same simulation on cluster compute](simulate-on-the-cluster.qmd) picks the same simulation up unchanged.

```{r}
#| label: setup
#| message: false
library(purrr)
library(tibble)
library(dplyr)
library(parallel)
```

## Is your problem this shape?

The shape is one independent unit per site, per polygon, per resample or per replicate.

The test: if replicate 200 does not need anything replicate 199 produced, the work is this shape, and everything below applies. Almost all simulation, bootstrapping and resampling is. If your replicates do depend on each other, in a Markov chain for instance, none of this applies and the page stops being useful to you.

That property has a name, **embarrassingly parallel**, and it is worth knowing because it is the term every parallel-computing document you meet next will use for it, including the vendor's. It is also the reason the escalation on this page works at all: independent units can be handed out in any order, to anything, without coordination.

Recognising the shape matters more than matching your problem to an example. You will know your own problem faster than you will find it in a list.

## The simulation this page runs

The examples below run one concrete simulation, because the shape is easier to see with a reason attached than without.

The question is how many days a year a river's flow sits below a low-flow threshold, and how uncertain that figure is given a finite record. The threshold is the station's own tenth percentile, so it is a low flow for that river rather than a number carried in from a bigger one. A single fit gives the central figure; what it does not give is how much to trust it, because the mean and standard deviation behind it were themselves estimated from a limited record.

So each draw perturbs the fitted parameters by their own sampling error and asks what fraction of a year falls below the threshold under the perturbed values. The spread across draws is the interval, and a station with three years of record gets a wider one than a station with ten.

The unit of work is one station. That is the shape that matters here: a few hundred independent units, each cheap, which is what most of this work actually looks like.

```{r}
#| label: units
set.seed(1)

n_stations <- 400

stations <- tibble(
  station_id = sprintf("S%04d", seq_len(n_stations)),
  n_days = sample(1000:3650, n_stations, replace = TRUE),
  meanlog = rnorm(n_stations, 0.5, 0.4),
  sdlog = runif(n_stations, 0.6, 1.1)
)

head(stations, 3)
```

The parameters are simulated here so the page stands alone and needs no connection. In your own work they would come from a server-side summary, which is what [Work with a table that will not fit](big-table.qmd) is about: the table shrinks to one row per station before any of this starts.

## Running them serially

Start here and keep it. The serial version is the one you can debug, and it is what every faster version has to agree with.

```{r}
#| label: serial
simulate_station <- function(station) {
  set.seed(station$n_days)

  se_meanlog <- station$sdlog / sqrt(station$n_days)
  se_sdlog <- station$sdlog / sqrt(2 * (station$n_days - 1))
  threshold <- qlnorm(0.1, station$meanlog, station$sdlog)

  draws <- map_dbl(1:1000, \(k) {
    m <- rnorm(1, station$meanlog, se_meanlog)
    s <- rnorm(1, station$sdlog, se_sdlog)
    plnorm(threshold, m, s)
  })

  tibble(
    station_id = station$station_id,
    days_below = 365 * mean(draws),
    lower = 365 * quantile(draws, 0.025, names = FALSE),
    upper = 365 * quantile(draws, 0.975, names = FALSE)
  )
}

results <- stations |>
  split(seq_len(nrow(stations))) |>
  map(simulate_station) |>
  list_rbind()

head(results, 3)
```

The central figure is near 36.5 days for every station, which is arithmetic rather than a finding: the threshold is each station's own tenth percentile, so a tenth of a year is the answer by construction. The number carrying information is the width of the interval.

## Running them on your own cores

One line changes:

```{r}
#| label: parallel
#| message: false
#| warning: false
library(future)
library(furrr)

plan(multicore, workers = 2)

results_parallel <- stations |>
  split(seq_len(nrow(stations))) |>
  future_map(simulate_station, .options = furrr_options(seed = NULL)) |>
  list_rbind()

identical(results, results_parallel)
```

That is the whole argument for trying your own cores first: `map()` becomes `future_map()`, you add a plan, and nothing else about the code moves. Given how little it costs to try, try it before you consider anything involving the cluster.

The comparison above is the point of running it. Same answer, exactly, which is what tells you the parallel version is doing the serial version's work and not something subtly different.

What it buys you depends on how many cores you actually have, which is [Choosing a worker count](#choosing-a-worker-count) below, and on how much work each unit does. This page states no timing, because the machine that builds it has one core and any number from it would describe this container rather than your session. Time your own serial run and multiply: that arithmetic is worth more than a figure from someone else's machine.

Seeding needs a decision, and which way you go depends on where the seed lives.

`simulate_station()` seeds itself from a column of its own input, so every station's draws are determined by the station rather than by the order the work happens to run in. That is why `seed = NULL` is right here and why the two runs match exactly: nothing in the parallel machinery is allowed to touch the streams.

If instead your function does *not* seed itself, you need `furrr_options(seed = TRUE)`, which gives each worker an independent, reproducible stream. Without it the workers can draw the same numbers, and `furrr` warns you about exactly that. What you must not do is both: `seed = TRUE` overrides a seed set inside the function, and then the serial and parallel runs disagree for a reason that looks like a bug in your simulation.

## Memory when forking

Forking to run the repeats in parallel does not raise peak memory. It lowers it.

```{r}
#| label: memory
peak_mb <- function(expr) {
  invisible(gc(reset = TRUE))
  force(expr)
  g <- gc()
  round(sum(g[, "max used"] * c(56, 8)) / 1e6, 1)
}

first_8 <- stations |> head(8) |> split(1:8)

tibble(
  run = c("serial", "forked, 2 workers"),
  peak_mb = c(
    peak_mb(map(first_8, simulate_station) |> list_rbind()),
    peak_mb(mclapply(first_8, simulate_station, mc.cores = 2) |> list_rbind())
  )
)
```

Each worker holds only its own slice and returns a small summary, so the parent never has every intermediate in memory at once. This is the opposite of the big-table case, where pulling more rows into the session is exactly what hurts, and it is worth knowing because it removes a worry that would otherwise stop you parallelising.

## `multicore` versus `multisession`

`multisession` starts fresh R processes and copies what they need. `multicore` forks, so each child starts as a copy-on-write view of the parent and pays nothing for data it only reads.

That depends on forking. Under `multisession` each worker gets its own full copy, and the peak goes up rather than down.

`multisession` is the plan people reach for by default, partly because it is what works on Windows. You are on Linux here, where forking is available, so prefer `multicore` and know that the two are not interchangeable for this purpose.

## Choosing a worker count

Do not agonise over the number. The penalty for a wrong count is mild in both directions, and the fix is one number and a re-run.

Too few workers and you leave cores idle, which costs you some time. Too many and they contend, which costs you a little more. Neither corrupts anything. A reasonable guess, run it, adjust if it matters.

One caution specific to containers: `detectCores()` reports the machine's cores, not the share your container is allowed. The two do not always agree, and the direction that hurts is asking for more workers than your quota permits.

## Packages inside the worker

A forked worker inherits your session's loaded packages, so a local fork is fine. A worker on cluster compute is a different R process on a different machine, and it needs `sf` installed there.

Check before you build a pipeline on it, then see [Packages](../ref/packages.qmd). This depends on your administrator having built a library volume. If they have not, this section does not apply to you, and that is a conversation rather than something to solve in R.

## Bootstrapping and resampling

A bootstrap replicate does not need the previous one, so everything on this page applies unchanged.

## Tuning across resamples

`tune` already parallelises over resamples through the same `future` plan. The plan you set earlier applies, and there is no separate step to take.

So if you have set `plan(multicore, workers = 2)` and then call `tune_grid()`, the resamples spread across those workers. You do not pass anything to the tuning call, and you do not need a `tune`-specific option.

The corollary is worth holding on to: if you have not set a plan, tuning runs serially and nothing tells you so.

## Is the cluster worth it?

Often not. If your repeats finish on your own cores in an afternoon, that is the better route. Reaching for distribution costs setup time, adds a serialisation step, and introduces a second environment whose package library you do not control. None of that buys anything if the work already fits.

The arithmetic is usually enough to decide. Time one replicate, multiply by how many you need, divide by the cores you have. If the answer is hours, you are done thinking. If the answer is days, read on.

Sometimes the honest answer is that your laptop is the right machine. If the data fits and nobody dictates where the computation happens, local R is faster to set up and better for your science.

## What changes on the cluster

Mostly the signature, and it is worth knowing before you decide.

`spark_apply()` hands your function a data frame holding many rows and expects a data frame back, so `simulate_station()` above becomes the body of a loop rather than the thing that is called. The simulation inside it does not change.

Three smaller things do. A forked worker inherits your loaded packages and a cluster worker does not. `furrr` is absent from the cluster runtime, so the far side uses base `parallel` instead. And the partition count replaces `workers =` as the dial.

[Run the same simulation on cluster compute](simulate-on-the-cluster.qmd) does that reshaping on this page's own simulation, and confirms the work left your session.

## Next

If a package is missing where the repeats run, [Packages](../ref/packages.qmd). If something broke, [When it breaks](../ref/when-it-breaks.qmd) is organised by what you saw.
