---
title: "Work with rasters, using `terra`"
subtitle: "Why gridded data behaves differently from polygons here, and what your two options are"
---

## `terra` has no server-side equivalent

There is no server-side path for gridded data. The native spatial functions on the warehouse are `ST_` functions, and every one of them takes a geometry: a point, a line, a polygon. None of them takes a grid, a band or a cell, so there is nothing for a raster operation to be pushed into.

If you arrived here from [Join and transform polygons](polygons.qmd), the answer there was that the spatial work can stay on the server and only a small result needs to cross. That answer does not generalise, and it is worth being clear about why. The `ST_` functions cover vector geometry, so a polygon intersection or an area has somewhere to run. A raster has no such counterpart, and it is not a gap that someone is about to fill: it is what the function family is for. Assuming otherwise costs you an afternoon of looking for the function that does not exist.

## So you have two options

Bring the data to R, or send R to the data. There is no third option, so this is a smaller decision than it looks, and most of this page is about making the second one work rather than about choosing between them.

Bringing the data to R is the ordinary case, and it is the right answer more often than the platform framing suggests. If the raster fits in your session, read it and work in `terra` exactly as you would anywhere else. Nothing on this page applies.

Sending R to the data means running your `terra` code inside a worker on the cluster, where the constraint is not the arithmetic but getting the pixels there. That is where the rest of the page goes.

## Check whether `terra` loads where your code runs, before anything else

Two lines, run in the context you are asking about:

```{r}
#| label: mask
#| include: false
source(here::here("R/setup.R"))
wq_mask_output()
```

```{r}
#| label: gdal
library(terra)
terra::gdal()
```

A GDAL version number means the system libraries are there and `terra` found them, and you can stop here. Most readers can. If either line fails, see [Packages](../ref/packages.qmd), and expect the cause to be a missing system library rather than a missing R package: `terra` needs GDAL, GEOS and PROJ underneath it, and those are not something you can add from your own session.

Run it twice, though: once in your own session and once inside a worker. They are different machines with different libraries, and a `terra` that loads locally tells you nothing at all about the worker. That is the most common failure on this page and the cheapest to rule out, which is why it comes before anything else.

## If you send R to the data

A disk-backed raster serialises its file path, not its pixels. So when you close over a `SpatRaster` and send the function to a worker, what arrives is a reference to a file that the worker has to be able to open. If it cannot, the pixels are not there.

This is the fact that decides whether the option works at all, and it is unintuitive, because in your session the object looks entirely self-contained. It prints its extent, its resolution, its CRS and its layer names, and it will hand you values on request. The pixels are not in it. Compare what the object serialises to against the file it came from:

```{r}
#| label: serialised
library(terra)

f <- system.file("ex/elev.tif", package = "terra")
r <- rast(f)

c(
  file_bytes = file.size(f),
  serialised_bytes = length(serialize(r, NULL)),
  cells = ncell(r),
  in_memory = inMemory(r)
)
```

Eight and a half thousand cells, and the whole serialised object is `r length(serialize(r, NULL))` bytes. It is smaller than the file it describes, and there is no arrangement of `r format(ncell(r), big.mark = ",")` values that fits in it. `in_memory` is 0, which is the same fact said directly.

What the object holds instead is the source, and `terra` will tell you so:

```{r}
#| label: sources
basename(sources(r))
```

The consequence is that the failure arrives in the worker as a missing file, which reads like a permissions problem rather than a serialisation one. You can see the shape of it locally, without a cluster, by writing a raster out, serialising it, then taking the file away before you unserialise. That is exactly what a worker experiences:

```{r}
#| label: missing-file
#| error: true
copy <- file.path(tempdir(), "elev-copy.tif")
writeRaster(r, copy, overwrite = TRUE)

b <- serialize(rast(copy), NULL)
file.remove(copy)

values(unserialize(b))[1:3]
```

That is the message a worker gives you, and it is worth reading closely. It names a path, and the path is one from the machine that built the object rather than the machine that failed to open it. Nothing in it mentions serialisation, so the natural reading is that the worker cannot see a file it should be able to see, which sends you to look at permissions and mounts. The object never carried the pixels in the first place.

So the file has to be reachable from the worker, which means a Unity Catalog volume rather than your own filesystem. A path under your home directory, or a `tempfile()` you wrote in your session, is a path the worker cannot open, however correct it looks in the closure.

The alternative is to stop the object being disk-backed before you send it, by reading the pixels into memory so that they are genuinely part of what gets serialised. That works, and it moves the constraint rather than removing it: what you can put in a closure is a small-object budget, not a raster-sized one. [Sending things to where your code runs](../ref/sending-things.qmd) has where that limit sits.

## Reading `/Volumes/...` from inside a worker

The worker reads a volume path as an ordinary local path, and that is the good news on this page. There is no special client, no download step and no API call: `rast("/Volumes/catalog/schema/volume/dem.tif")` inside the worker function opens the file the way it would open anything else, and GDAL is satisfied.

Which means the working shape is: write the raster to a volume once, then have the worker function open it by path. The path is a string, so it costs nothing to close over, and the pixels never travel through the closure at all.

The mechanics of getting the file onto the volume in the first place are in [Sending things to where your code runs](../ref/sending-things.qmd).

## What this rules out, stated plainly

Some raster workflows have no good answer here, and it is better to know which before you start.

Anything that needs the whole grid at once on the server has nowhere to run. There is no server-side resample, no zonal statistic across a raster and a polygon layer, no map algebra between two rasters in a query. Every one of those has to happen in R, either in your session or in a worker, and the platform contributes nothing except the machine it runs on.

Anything that would move a large raster repeatedly is paying for the move rather than the work. If the analysis is one pass over one grid, the transfer is most of the cost and distributing it buys you nothing.

And the honest case: if the raster fits on your laptop and nobody is telling you where the work has to happen, do it locally. Local R is faster to set up, it is easier to iterate on, and there is no server-side raster capability you are giving up, because there is none to give up. Sending R to the data earns its complexity when the raster is genuinely too large for your session, or when it already lives on a volume and moving it out is the expensive part. Neither of those is the common case.

## Next

[Sending things to where your code runs](../ref/sending-things.qmd).
