---
title: "Read my files from a volume"
subtitle: "Getting a CSV, a workbook or a shapefile out of Unity Catalog storage and into your session"
---

## `/Volumes/...` is not a path on your machine

If your data is a file that somebody else put in a volume, this is the page you need first, and this is the sentence that saves you the afternoon: **that `/Volumes/...` path is a filesystem path inside a Databricks cluster, and nowhere else.**

From your own R session it does not exist. So the obvious first move fails:

```{r}
#| label: mask
#| include: false
# This site is public, so no printed output may carry the catalog this page
# happens to run against. Only what is *shown* is masked; every call below
# runs against the real volume. See R/setup.R.
source(here::here("R/setup.R"))
wq_mask_output()
```

```{r}
#| label: setup
#| message: false
library(dplyr)
library(readr)
library(arrow)
library(sf)
library(brickster)
library(glue)
library(cli)

catalog <- Sys.getenv("DATABRICKS_CATALOG")
schema <- Sys.getenv("DATABRICKS_SCHEMA")

volume_root <- glue("/Volumes/{catalog}/{schema}/raw")
```

```{r}
#| label: nopath
#| error: true
read_csv(glue("{volume_root}/sample/stations.csv"))
```

Nothing is misconfigured and you have not mistyped the path. The path is real, the file is there, and your session has no way to see either. The message is the same one you would get for a typo, which is why this costs people time rather than seconds.

Everything below follows from that one fact. The file has to travel to you before R can open it.

## Fetch the file, then open it the way you always would

`brickster::db_volume_read()` brings a file down to a local path of your choosing. From there nothing is Databricks-specific: you open it with whatever you already use.

```{r}
#| label: readcsv-fetch
#| results: hide
local_csv <- file.path(tempdir(), "stations.csv")

db_volume_read(glue("{volume_root}/sample/stations.csv"), destination = local_csv)
```

```{r}
#| label: readcsv
stations <- read_csv(local_csv, show_col_types = FALSE)

stations |>
  select(station_id, station_name, easting, northing, river_name) |>
  head(3)
```

This is the default route, and it is the one to reach for unless you have a specific reason not to. It needs neither a cluster nor a warehouse, so there is no wait for compute to start. It moves bytes rather than parsing them, which is why the format does not matter: the same two lines work for Parquet with `arrow::read_parquet()`, for a workbook with `readxl::read_excel()`, and for anything else you can open locally.

```{r}
#| label: readparquet-fetch
#| results: hide
local_pq <- file.path(tempdir(), "stations.parquet")

db_volume_read(
  glue("{volume_root}/hydrology_stations/data.parquet"),
  destination = local_pq
)
```

```{r}
#| label: readparquet
read_parquet(local_pq) |>
  dim()
```

Writing goes the other way with `db_volume_write()`, which does take an `overwrite` argument where `db_volume_read()` does not. The documented ceiling is 5 GiB per file, which is the same figure [Sending things to where your code runs](../ref/sending-things.qmd) quotes.

To see what is actually in a directory, `db_volume_list()`:

```{r}
#| label: list
db_volume_list(glue("{volume_root}/sample"))$contents |>
  purrr::map_chr(\(item) basename(item$path))
```

Do not reach for `db_volume_file_exists()` to guard one of these calls. It is documented as returning `TRUE` or `FALSE`, and on a path that does not exist it throws `HTTP 404 Not Found` instead, so it fails on exactly the branch you wrote it to test. Listing the parent directory, as above, answers the question without the surprise.

## A shapefile is four files, so fetch the directory

This is the sharpest trap on the page, because one half of it is silent.

A shapefile is not a file. It is at least four, sitting beside each other: `.shp` holds the geometry, `.shx` the index, `.dbf` the attributes, and `.prj` the coordinate reference system. Fetch the directory rather than the parts, with `db_volume_download_dir()`:

```{r}
#| label: shapefile-fetch
#| results: hide
shp_dir <- file.path(tempdir(), "stations_shp")

db_volume_download_dir(glue("{volume_root}/sample/stations_shp"), shp_dir)
```

```{r}
#| label: shapefile
list.files(shp_dir)
```

```{r}
#| label: shaperead
stations_sf <- st_read(file.path(shp_dir, "stations.shp"), quiet = TRUE)

cli_alert_info(
  "{nrow(stations_sf)} features, EPSG:{st_crs(stations_sf)$epsg}"
)
```

The argument for the directory call rather than four `db_volume_read()` calls is the asymmetry in how the two possible mistakes behave.

Forget the `.shx` and you find out immediately: `st_read()` fails, and the GDAL error names the missing file. Annoying, and self-correcting.

Forget the `.prj` and **nothing tells you anything**. The read succeeds, you get every feature, and the coordinate reference system is quietly gone:

```{r}
#| label: noprj
#| results: hide
partial_dir <- file.path(tempdir(), "stations_noprj")
dir.create(partial_dir, showWarnings = FALSE)

# The three parts somebody fetching by hand would think of.
for (ext in c("shp", "shx", "dbf")) {
  db_volume_read(
    glue("{volume_root}/sample/stations_shp/stations.{ext}"),
    destination = file.path(partial_dir, glue("stations.{ext}"))
  )
}
```

```{r}
#| label: noprj-read
partial <- st_read(file.path(partial_dir, "stations.shp"), quiet = TRUE)

cli_alert_info("{nrow(partial)} features, CRS: {format(st_crs(partial)$epsg)}")
```

Same feature count, no error, no warning, and `st_crs()` is `NA`. If you are working in EPSG:27700 and the projection has silently become unknown, every distance and area you compute afterwards is suspect, and the first thing that will tell you is a map of Britain in the wrong shape. One `db_volume_download_dir()` call cannot make this mistake, which is the whole reason to prefer it.

Worth knowing for your own outputs: a GeoPackage is a single file, so the entire class of problem disappears. If you control the format, write `.gpkg` rather than `.shp` and the four-part bookkeeping goes away.

## When to leave the file on the server instead

Everything above brings the whole file to you. That is right until the file is large and you only want part of it, at which point you are downloading gigabytes to throw most of them away.

For Parquet or CSV, `read_files()` reads the volume path server-side, so the filtering happens before anything crosses the wire. It composes with `dbplyr` through `tbl(con, sql(...))`, so your pipeline still looks like the pipelines everywhere else on this site:

```{r}
#| label: readfiles
#| message: false
library(DBI)
library(dbplyr)

con <- dbConnect(
  odbc::databricks(),
  httpPath = Sys.getenv("DATABRICKS_PATH"),
  DefaultStringColumnLength = 65535
)

files_query <- glue(
  "SELECT * FROM read_files(
     '{volume_root}/hydrology_stations/data.parquet',
     format => 'parquet'
   )"
)

tbl(con, sql(files_query)) |>
  filter(!is.na(catchment_area_km2)) |>
  summarise(
    stations = n(),
    mean_area_km2 = mean(catchment_area_km2, na.rm = TRUE)
  ) |>
  collect()
```

This needs a SQL warehouse, which is the trade: you get server-side filtering and you give up the no-compute-required property of the route above.

One behaviour to know before you point it at a directory. `read_files()` reads **every file in a directory regardless of extension**, so a directory holding a CSV next to a Parquet file fails rather than reading either. Give it a glob when the directory is mixed:

```sql
read_files('/Volumes/<catalog>/<schema>/raw/some_dir/*.parquet',
           format => 'parquet')
```

`**/*.parquet` recurses into subdirectories if you need that.

## Excel

Download it and use `readxl::read_excel()`. That is the whole answer.

Do not route a workbook through the server: `read_files()` does not support xlsx, and it was never the right tool for it. Because `db_volume_read()` moves bytes without parsing them, a workbook arrives intact with every sheet, and `readxl` reads it exactly as it would a workbook on your laptop.

## Which route to use

| Your situation | Route |
|----|----|
| Any format, any size up to 5 GiB, no compute running | `db_volume_read()`, then open locally |
| A shapefile, or any multi-part format | `db_volume_download_dir()` on the containing directory |
| A workbook | `db_volume_read()`, then `readxl::read_excel()` |
| Parquet or CSV, large, and you want a subset | `read_files()` in `tbl(con, sql(...))` |
| Writing a file back | `db_volume_write()`, or `db_volume_upload_dir()` for a directory |

Reading and writing files inside a volume needs `READ VOLUME` and `WRITE VOLUME` on it. Neither is `CREATE VOLUME`, which you need only to make a new volume and which nothing on this page requires. If the volume already exists and you can list it, you can do everything here.

## Next

If your data is a governed table rather than a file, [connect your R session to the data](connect.qmd) is the page you want instead. If the file you just fetched is geometry, [join and transform polygons](polygons.qmd) picks it up from there. To send a file the other way, to where distributed code runs, see [Sending things to where your code runs](../ref/sending-things.qmd).
